Reproductive experience promotes permanent body growth independently of growth hormone

Version 1

Abstract

Pregnancy leads to many adaptations in the maternal body, most of which are reversible. However, reproductive experience can also result in permanent effects. Here, we investigated how pregnancy influences the somatotrophic system and the lasting effects of reproductive experience on the maternal organism. Reproductive experience induced a pronounced increase in lean body mass and longitudinal growth in both wild-type and growth hormone (GH)-deficient mice compared with age-matched virgins. Body growth was primarily observed during the first pregnancy, whereas a second gestation was mostly associated with increased adiposity. Data from a cohort of women with isolated GH deficiency (IGHD) caused by a loss-of-function mutation in the GHRHR gene revealed that nulliparous women were 7 cm shorter than those with one or more pregnancies. Increased GH secretion was observed in pregnant wild-type mice but not in pregnant GHRHR-deficient mice. Pregnancy-induced body growth is preserved despite disruption of GH-, ghrelin-, and estrogen-related signaling pathways. In conclusion, reproductive experience induces permanent changes in the maternal organism, promoting body growth in models that allow this response. Pregnancy-induced body growth appears to be independent of GH action. These findings underscore the need for further studies to investigate the long-lasting consequences of reproductive experience in females.

Introduction

Pregnancy is a unique state in which numerous physiological adaptations are necessary in the maternal organism to allow proper fetal growth and to prepare for the lactation period (Augustine et al., 2008). Among the diverse maternal adaptations, metabolic changes are particularly prominent, including alterations in appetite and glucose homeostasis, as well as increased adiposity (Augustine et al., 2008; Ladyman et al., 2010; Zampieri et al., 2015; Teixeira et al., 2019a). Pregnancy also affects neuroplasticity and causes behavioral and emotional alterations (Larsen and Grattan, 2012; Georgescu et al., 2021). Most of these changes are reversible. However, pregnancy may lead to long-term or permanent consequences. For example, prior pregnancy experience reduces the latency to display maternal behavior in rats (Bridges, 1978), indicating long-lasting behavioral adaptations strongly shaped by prior hormonal exposure. Primiparous mice show increased body weight and reduced ambulatory activity compared with age-matched virgin animals, even several months after last contact with the pups (Ladyman et al., 2018; Teixeira et al., 2019b; Ladyman et al., 2021). Body weight retention is frequently observed after pregnancy, especially for women who experienced excessive gestational weight gain (Scholl et al., 1995; Rooney et al., 2005). Thus, pregnancy may represent a risk factor for obesity in women (Amorim et al., 2007).

Pregnancy-induced adaptations are thought to be largely driven by the marked rise in circulating hormones during gestation (Augustine et al., 2008), including progesterone and estrogens, as well as placentally derived analogues of pituitary hormones, such as chorionic gonadotropin and placental lactogens (also referred to as chorionic somatomammotropins). However, important differences are observed between humans and rodents. Humans, but not rodents, possess the GH2 gene, which encodes the placental growth hormone (GH) variant, whose production progressively replaces pituitary GH (derived from the GH1 gene) secretion during pregnancy (Liao et al., 2018). In contrast, rodents maintain pituitary GH secretion throughout gestation (Gatford et al., 2017; Liao et al., 2018). Additionally, human GH, both pituitary and placental variants, as well as chorionic somatomammotropins, can activate both the GH receptor (GHR) and the prolactin receptor (PrlR), explaining their lactogenic effects. In contrast, murine GH activates only the GHR (Bartke and Kopchick, 2015).

Despite species-specific differences, it is evident that hormonal exposure during pregnancy not only transiently shapes maternal physiology but may also exert lasting effects on health across the lifespan. However, the chronic consequences of reproductive experience remain poorly understood. Given that most women will experience at least one pregnancy during their lifetime, elucidating how reproductive experience influences long-term physiology and health is of substantial biological and clinical importance. Furthermore, historically, females have been underrepresented in studies (Clayton and Collins, 2014; Morselli et al., 2016; Lee, 2018). Therefore, the goal of this study is to explore how pregnancy impacts the somatotrophic system and to assess the lasting effects on the maternal organism.

Results

Pregnancy promotes permanent body growth in wild-type and GH-deficient mice

To investigate the consequences of reproductive experience, body weight and composition of C57BL/6J wild-type (WT) mice were initially measured before mating. Pregnant mice gave birth in individual cages and were maintained with their litters for 3 weeks. Two weeks after weaning (the recovery period from lactation), the females were reevaluated to determine changes relative to baseline. The control group consisted of age-matched virgin females. These assessments were also carried out in Ghrhrlit/lit mice, which carry a null mutation in the gene encoding the GHRH receptor (GHRHR) and are therefore GH-deficient and dwarf. Reproductive experience caused a robust increase in body mass (Figure 1A) and lean mass (Figure 1B) in both control and Ghrhrlit/lit mice compared with virgin groups. No changes in fat mass were observed between groups (Figure 1C). A new evaluation was performed 2 weeks after the second pregnancy and lactation cycle. Body length measurements revealed a significant increase in multiparous WT and Ghrhrlit/lit mice compared with age-matched virgin animals (Figure 1D-E). In addition, multiparous mice showed higher body and lean mass than virgin animals (Figure 1F-G). Despite pregnancy-induced growth, Ghrhrlit/lit mice remained smaller than WT females (Figure 1D-G). Fat mass increased after 2 pregnancies in both WT and Ghrhrlit/litmice compared with their respective age-matched virgin groups (Figure 1H). Multiparous WT and Ghrhrlit/lit mice also had increased liver (Figure 1I) and kidney (data not shown) masses compared with virgin groups. Heart and brain masses increased only in multiparous Ghrhrlit/lit mice compared with virgin Ghrhrlit/lit females, whereas these tissues did not change in WT mice (data not shown). Thus, pregnancy promotes body growth in WT and GH-deficient females.

Pregnancy promotes permanent body growth in wild-type and GH-deficient mice.

(A-C) Differences relative to baseline in body, lean, and fat mass in virgin WT (n = 15), primiparous wild-type (WT; n = 13), virgin Ghrhrlit/lit (n = 6), and primiparous Ghrhrlit/lit(n = 13). (D) Representative image showing differences in longitudinal growth between virgin and multiparous WT and Ghrhrlit/lit mice. (E-I) Body length, body weight, lean mass, fat mass, and liver mass in multiparous WT (n = 13) and Ghrhrlit/lit (n = 13) mice, relative to age-matched virgin WT (n = 15) and Ghrhrlit/lit(n = 12) mice. *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Growth is primarily induced during the first pregnancy

The following experiment tested whether a second pregnancy could also increase parameters indicative of body growth. Unlike the body weight and lean mass gains observed after the first pregnancy, WT mice subjected to a second pregnancy cycle showed no increases in body or lean mass compared with age-matched WT females (Figure 2A-B). Because females undergoing a second pregnancy are inherently older than during their first gestation, we assessed whether a first pregnancy occurring at a comparable age to the second would influence body weight. WT females that had a late first pregnancy (age-matched first pregnancy group) showed increased body and lean mass gain compared with age-matched virgin and second pregnancy groups (Figure 2A-B). Notably, the gains in body and lean mass induced by the first pregnancy were similar among the youngest and oldest females compared with their respective age-matched groups of virgin females (Figure 2A-B). No differences in fat mass gain were observed between the groups (data not shown). In contrast to WT mice, the second pregnancy resulted in a significant increase in body and lean mass in Ghrhrlit/lit females compared with age-matched virgin animals (Figure 2C-D). However, the magnitude of the increase in the second pregnancy was significantly lower than in the first (Figure 2C-D), highlighting that growth occurs primarily during the first pregnancy.

Growth is primarily induced during the first pregnancy.

(A-B) Changes in body weight and lean mass in wild-type (WT) virgins (n = 15), after the first pregnancy (n = 13), in virgins matched for age at the second pregnancy (n = 15), after the second pregnancy (n = 13), and after a late first pregnancy (n = 8). (C-D) Changes in body weight and lean mass in Ghrhrlit/lit virgins (n = 6), after the first pregnancy (n = 13), in virgins matched for age at the second pregnancy (n = 12), and after the second pregnancy (n = 13). *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using unpaired two-tailed Student’s t-test or one-way ANOVA and Newman-Keuls multiple comparisons test.

Case reports of Itabaianinha women with isolated GH deficiency suggest post-pregnancy growth

Growth plate physiology differs between humans and rodents: growth plates remain open after sexual maturation in rodents, whereas they fuse at the end of puberty in humans, thereby terminating longitudinal bone growth (Emons et al., 2011). Although pregnancy-induced longitudinal growth is not observed in normal adult women, we speculate that a slight pregnancy-induced growth may happen in certain special cases. To provide proof-of-concept evidence that gestational factors permanently influence the maternal organism in humans, we studied a cohort of individuals with severe isolated GH deficiency (IGHD) caused by a homozygous loss-of-function mutation (c.57+1G→A) in the GHRHR gene. These individuals are from the Brazilian city of Itabaianinha (Salvatori et al., 1999; Aguiar-Oliveira and Salvatori, 2021), and their mutation replicates effects seen in Ghrhrlit/lit mice (Aguiar-Oliveira and Bartke, 2019).

Using data from this rare cohort, we compared the height of 8 nulliparous patients and 9 patients with one or more pregnancies, both groups with IGHD without lifetime GH replacement therapy. The average number of children per woman was 1.8 children (Aguiar-Oliveira and Salvatori, 2021). Nulliparous women were 7 cm shorter than those with one or more pregnancies, 1.13 ± 0.04 m vs 1.20 ± 0.05 m (P = 0.009; Figure 3A). Moreover, the three shortest women, with heights of 1.07, 1.08, and 1.11 m, were nulliparous. Anecdotal reports from 6 IGHD cases also suggest residual growth in 4 women after pregnancy. This information was obtained through face-to-face personal interviews with one of the researchers (M.H. A.-O.). Patient 1 reported clear and consistent signs of somatic growth following two pregnancies. She is married to an IGHD individual, and both of her sons (currently 30 and 33 years old) are affected, as expected for this genetic background. According to documented measurements, her height increased from 1.16 m before pregnancy to 1.20 m after her two pregnancies. She also reported an increase in shoe size from 25 to 28 (Brazilian size) and changes in ring size, as a ring previously worn on the fourth finger no longer fit after childbirth. Together, these findings are consistent with clinically perceptible post-pregnancy somatic growth.

Evidence of post-pregnancy growth in women with isolated GH deficiency.

(A) In a cohort of women with isolated GH deficiency caused by a loss-of-function mutation in the GHRHR gene, nulliparous women (n = 8) were shorter than those with one or more pregnancies (n = 9). **, P = 0.009 (unpaired two-tailed Student’s t-test). (B) Comparison of foot size between two sisters with IGHD, one nulliparous and another primiparous. Image of the feet of two sisters with severe isolated GH deficiency caused by a homozygous loss-of-function mutation in the GHRHR gene. In addition to a larger shoe size, patient 2, who had a heterozygous son 9 years ago, is also taller than her nulliparous sister.

Three additional women with IGHD also reported body changes suggestive of growth after pregnancy, though less noticeable than those of patient 1. Patient 2 (43 years old) had a documented increase in height from 1.26 m before pregnancy to 1.27 m at the present. She has one heterozygous son (8.8 years old). She reported an increase in shoe size from 29 to 30, and her graduation ring no longer fits; however, she also gained weight, which may have contributed to the ring size changes. Her feet are larger than her sister’s; she is 44 years old, 1.21 m tall, shoe size 27, and nulliparous (Figure 3B). Patient 3, treated with GH in childhood and early adolescence, was 1.44 m tall before pregnancy. She had a heterozygous boy 13 years ago; now she is 36 years old and 1.47 m tall. Patient 4, currently 28 years old, was treated with GH in childhood and early adolescence. She had a heterozygous girl 3 years ago. Her current height is 1.42 m, but before pregnancy, she was 1.40 m tall. In contrast, 2 sisters with IGHD reported no signs of pregnancy-induced growth. Patient 5, 1.24 m tall, aged 56, has a daughter, aged 35, and two sons, aged 30 and 29, each heterozygous. Patient 6, 1.27 m tall, aged 68, has two heterozygous daughters, aged 46 and 44, respectively. However, no information about their heights before pregnancy is available. Despite these two cases, our data indicate a slight post-pregnancy growth in women with severe IGHD.

Increased GH secretion in pregnant WT mice but not in Ghrhrlit/lit females

Because GH is the major circulating factor that induces longitudinal growth in mammals (Zhou et al., 1997; Dehkhoda et al., 2018), we investigated whether WT and Ghrhrlit/lit mice exhibit alterations in GH secretion during pregnancy (gestational age 14-17 days). Confirming previous studies (Gatford et al., 2017; Kaur et al., 2020; Wasinski et al., 2022), pregnancy increased GH secretion in WT mice compared with virgin animals (Figure 4A, C). This increase was associated with higher GH pulse frequency and elevations in both pulsatile and basal secretion (Figure 4C-G). The contribution of basal to total GH secretion increased in pregnant WT mice (Figure 4H). As expected, non-pregnant Ghrhrlit/lit mice showed blunted GH secretion (Figure 4B). Importantly, pregnancy did not stimulate GH secretion in Ghrhrlit/lit mice (Figure 4C-H), indicating that GHRH signaling is required for pregnancy-induced increases in GH secretion in mice.

Increased GH secretion in pregnant WT mice but not in pregnant Ghrhrlit/lit mice.

(A-B) Representative examples of the pattern of GH secretion in two pregnant and virgin WT mice and in two pregnant and virgin Ghrhrlit/lit mice. (C-H). Total GH secretion, GH pulse frequency, pulsatile GH secretion, GH pulse amplitude, basal (non-pulsatile) GH secretion, and contribution of basal secretion to total GH secretion in virgin WT (n = 6), pregnant WT (n = 5), virgin Ghrhrlit/lit (n = 6), and pregnant Ghrhrlit/lit (n = 3) mice. The pregnant mice were at gestational ages 14-17 days. *, P < 0.05; **, P < 0.01; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Pregnancy differentially modulates hepatic GHR signaling in WT and Ghrhrlit/lit mice

GH action in the liver plays a key role in controlling growth because GHR signaling stimulates hepatic insulin-like growth factor 1 (IGF-1) secretion, particularly by activating the signal transducer and activator of transcription 5b (STAT5b) pathway (Udy et al., 1997; Teglund et al., 1998; Grimley et al., 1999). GH action in the liver is not only controlled by the pattern of GH secretion but also by hepatic GH sensitivity (Touvier et al., 2009; Yamamoto et al., 2013; de Sousa et al., 2025b). A reduced expression of STAT5b protein was observed in the liver of pregnant WT and Ghrhrlit/litmice (Figure 5A,C). Phosphorylation of STAT5 (pSTAT5) was also reduced in the liver of pregnant WT compared to virgin animals (Figure 5B-C). Hepatic pSTAT5 was reduced in Ghrhrlit/litmice, compared to virgin WT, and did not change in pregnant animals (Figure 5B-C). Serum IGF-1 and hepatic Igf1 mRNA levels were reduced in pregnant WT mice compared with virgin animals (Figure 5D-E). In contrast, pregnancy increased serum IGF-1 and hepatic Igf1 mRNA levels in Ghrhrlit/litmice (Figure 5D-E).

Pregnancy differentially modulates hepatic GHR signaling in WT and Ghrhrlit/litmice.

(A-C) Western blot analysis to determine the hepatic expression of STAT5b and pSTAT5 proteins in virgin WT (n = 6), pregnant WT (n = 6), virgin Ghrhrlit/lit (n = 6), and pregnant Ghrhrlit/lit (n = 6) mice. (D-K) Serum IGF-1 levels and hepatic mRNA expression in virgin WT (n = 11), pregnant WT (n = 10), virgin Ghrhrlit/lit (n = 8), and pregnant Ghrhrlit/lit (n = 7) mice. The pregnant mice were at gestational ages 14-17 days. *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Corroborating the results of the protein expression, pregnancy reduced Stat5b mRNA levels in both WT and Ghrhrlit/lit mice (Figure 5F). Conversely, hepatic Ghr and Prlr mRNA levels increased during pregnancy in WT and Ghrhrlit/lit mice (Figure 5G-H). Esr1 mRNA expression was reduced in Ghrhrlit/lit mice and was not affected by pregnancy (Figure 5I). Hepatic Socs2 and Cish RNA levels, transcripts that are indicative of GH action and regulate GH sensitivity (Greenhalgh and Alexander, 2004), were also reduced in Ghrhrlit/lit mice compared to WT animals (Figure 5J-K). Moreover, pregnancy increased Socs2 and Cish RNA expression only in Ghrhrlit/litmice (Figure 5J-K).

Altogether, although WT mice exhibit increased GH secretion and hepatic Ghr expression, molecular analyses do not indicate higher GH action in the liver. In contrast, Ghrhrlit/litmice showed evidence of greater GH action in the liver, including increased circulating and hepatic IGF-1 levels, as well as greater Socs2 and Cish expression.

Pregnancy-induced body growth is preserved despite disruption of GH-, ghrelin-, and estrogen-related signaling pathways

To gain further insights into the mechanisms driving post-pregnancy body growth, the effects of pregnancy were investigated across different mouse models. Mice carrying a GHR deletion in hepatocytes (AlbΔGHR) show blunted circulating IGF-1 levels and growth deficits despite higher GH secretion (List et al., 2014; Wasinski et al., 2023; de Sousa et al., 2025b). We observed that a first pregnancy caused a robust increase in body weight and lean mass in AlbΔGHR mice, which was comparable to that seen in control animals (Figure 6A-B). AlbΔGHR females also showed increased body weight after the second pregnancy, but this was explained by pregnancy-induced increases in fat mass (Figure 6C). Thus, GHR signaling in hepatocytes is not necessary to induce pregnancy-induced growth. Ghrelin is a GH secretagogue, and impaired ghrelin signaling may affect body growth (Peris-Sampedro et al., 2021; Punt et al., 2025).

Pregnancy-induced body growth is preserved despite disruption of GH-, ghrelin-, and estrogen-related signaling pathways.

(A-C) The effects of first and second pregnancies on body weight, lean mass, and fat mass in control virgin (n = 10), control 1st pregnancy (n = 7), control second pregnancy (n = 5), AlbΔGHR virgin (n = 17), AlbΔGHR 1st pregnancy (n = 9), and AlbΔGHR second pregnancy (n = 6). (D-E) Body weight and lean mass in control virgin (n = 5), control 1st pregnancy (n = 14), GHSR KO virgin (n = 4), and GHSR KO 1st pregnancy (n = 13). (F-H) Body weight, lean mass, and fat mass in control virgin (n = 7), control 1st pregnancy (n = 4), AlbΔERα virgin (n = 7), and AlbΔERα 1st pregnancy (n = 4). (I-K) The effects of first and second pregnancies on body weight, lean mass, and fat mass in control virgin (n = 11), control 1st pregnancy (n = 11), control second pregnancy (n = 6), NestinΔGHR virgin (n = 6), NestinΔGHR 1st pregnancy (n = 6), and NestinΔGHR second pregnancy (n = 3). *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

However, ghrelin receptor-deficient mice (GHSR KO) showed increases in body weight and lean mass similar to those of control animals (Figure 6D-E). Sex hormones also regulate growth, and estrogen receptor α (ERα) signaling activates the STAT5 pathway in the liver and IGF-1 synthesis (Venken et al., 2005). In addition, estradiol treatment stimulates growth in GHR-deficient mice (Venken et al., 2005), and estrogen response elements are found in the mouse Stat5b gene (Furigo et al., 2014). Nonetheless, hepatocyte-specific ERα ablation did not prevent pregnancy-induced increases in body weight, lean mass, and adiposity (Figure 6F-H). We also evaluated mice carrying brain GHR inactivation, as this is a model of gigantism caused by loss of GH negative feedback (Furigo et al., 2019; Teixeira et al., 2019a; Wasinski et al., 2020; de Sousa et al., 2025a). First pregnancy increased body, lean, and fat mass in NestinΔGHR females, as observed in control animals (Figure 6I-K).

Discussion

Pregnancy causes profound endocrine and metabolic changes that are conserved across mammals. The maternal organism undergoes significant hormonal shifts that not only temporarily alter the body but may also lead to lasting effects. Here, we demonstrated that reproductive experience resulted in increased body growth in WT mice, independent of GH. This growth was mainly observed after the first pregnancy, while subsequent reproductive experiences primarily increased body fat. In humans, puberty triggers the fusion of growth plates, leading to the end of longitudinal bone growth (Emons et al., 2011). Height gain is typically limited after menarche, averaging 6.6-8.94 cm depending on the studied cohort (Spear, 2002; Gardstedt-Berghog et al., 2024; Gaete et al., 2025; Xu et al., 2026). Previous studies indicate that pregnant women may experience a slight but significant decrease in height due to weight gain, vertebral compression, and increased lordosis (Scholl et al., 1993; Xu et al., 2026). Therefore, using height as a measure of post-pregnancy growth in humans may be unreliable, which may explain why normal gravidas do not show longitudinal growth. However, increased foot size has been reported after pregnancy (Wetz et al., 2006; Dunn et al., 2012). Despite the lack of evidence for post-pregnancy longitudinal growth in healthy women, the current results in mice suggest that pregnancy activates GH-independent growth pathways. In species with open growth plates, like rodents, this could result in actual longitudinal growth, whereas in humans, it may instead lead to tissue remodeling or physical changes that do not increase height but may have lasting effects that are still not well understood. Thus, the mouse model offers mechanistic insight into conserved pregnancy-related anabolic processes that may operate independently of the somatotropic axis.

Multiparous Ghrhrlit/lit mice showed significant gains in weight (+57%) and lean mass (+59%) after pregnancy compared with age-matched Ghrhrlit/lit virgins. In fact, their relative weight and lean mass gains were much higher than those observed in WT mice, which showed ∼13% more body weight and lean mass than WT virgins. Thus, GH-deficient mice are considerably more sensitive to pregnancy-induced growth than WT mice. The profound growth deficit in Ghrhrlit/litmice likely enhances their sensitivity to pregnancy-induced growth factors. Human data are in accordance with these findings as children with severe GH deficiency (GHD) show a higher response to GH replacement than patients with less severe or questionable GHD (Savage and Bang, 2012; Wit et al., 2019). Accordingly, patients exhibiting the lowest response during GH provocation tests showed the highest responses to GH replacement treatment (Donbaloglu et al., 2023). Regarding the capacity of women with IGHD to present post-pregnancy growth, residual spinal growth is frequently observed after skeletal maturity, particularly in the 1–2 years following the peak of puberty, often averaging just a few millimeters to 1 cm in height (Ghanem and Rizkallah, 2020). The increased responsiveness to growth factors, as well as the late menarche (±17 years) and, therefore, delayed bone maturation (Aguiar-Oliveira and Salvatori, 2021), can also explain why adult women with severe IGHD retained some growth capacity after pregnancy. Importantly, identifying adult women with untreated IGHD who subsequently experienced pregnancy is extremely challenging. GH replacement therapy is typically initiated early in life, and untreated individuals are rare. Therefore, although a limited number of cases were available for study, this sample represents a highly valuable and unique clinical resource.

Unlike WT females, Ghrhrlit/lit mice show no increase in GH secretion during pregnancy, maintaining suppressed levels similar to those of nulliparous Ghrhrlit/lit mice. Thus, these findings revealed that GHRH signaling is required for pregnancy-induced increases in GH secretion in mice. Remarkably, this result indicates that pregnancy-induced body growth is independent of GH levels. However, the expression of genes typically induced by GHR signaling, such as Igf1, Socs2, and Cish, was upregulated in the liver of pregnant Ghrhrlit/litmice compared with virgin animals. The increased circulating IGF-1 levels in Ghrhrlit/lit mice may have contributed to their greater relative growth during gestation. In contrast, pregnant WT mice showed decreased circulating IGF-1 levels and hepatic Igf1 mRNA expression compared to virgin animals. These results are consistent with another study that also observed decreased IGF-1 secretion and hepatic Igf1 mRNA expression in mid-pregnant mice (Travers et al., 1990). In humans, IGF-1 levels decrease during the first trimester of gestation (Caron et al., 2010; Persechini et al., 2015), but increase during mid- and late-pregnancy, when they correlate with placental GH levels (Caufriez et al., 1993). Altogether, pregnancy differentially modulates the hepatic expression of genes commonly induced by GHR signaling in WT and Ghrhrlit/lit mice.

In an attempt to identify specific hormonal factors involved in pregnancy-induced body growth, mice with hepatocyte-specific deletion of GHR or ERα were generated. Confirming previous findings that pregnancy-induced body growth is independent of GH, hepatic GHR deletion did not prevent increases in body weight and lean mass in primiparous and multiparous mice. Although estradiol treatment can stimulate growth in GHR-deficient mice and activate the STAT5 pathway and IGF-1 synthesis in the liver (Venken et al., 2005), hepatocyte-specific ERα KO mice also exhibited normal pregnancy-induced body growth. Similar findings were observed in GHSR-deficient mice, demonstrating that pregnancy-induced body growth is ghrelin-independent. Finally, a gigantism model (NestinΔGHR mice) was tested to determine whether pregnancy-induced growth can occur in mice that already have increased baseline lean body mass (+16%) compared with control animals. However, pregnancy also increased body and lean mass in NestinΔGHR mice, in a manner comparable to control animals, suggesting that this growth potential persists even in animals that begin gestation with greater somatic length.

Several questions arose from our findings. For example, why does body growth occur primarily after the first pregnancy, regardless of the age at which it occurs? We speculate that pregnancy hormones, particularly sex hormones, may modify the growth plate of long bones, limiting further growth in subsequent gestations. Future studies should investigate the impact of pregnancy on growth plate structure and function. Another interesting finding is that multiple gestations are associated with increased body adiposity in mice. Leptin resistance is a hallmark of pregnancy, favoring physiological increases in food intake and fat mass (Mounzih et al., 1998; Trujillo et al., 2011; Ladyman et al., 2012; Zampieri et al., 2015; Andreoli et al., 2019; Gustafson et al., 2019). If leptin resistance persists in the postpartum period, it may favor fat retention. However, previous studies indicate normal leptin sensitivity in female mice following pregnancy (Ladyman et al., 2018; Andreoli et al., 2019; Teixeira et al., 2019b).

It is also possible that each pregnancy may cause some degree of fat retention, so multiple pregnancies lead to a significant increase in body adiposity simply through an additive effect. Regardless of the mechanism that causes pregnancy-induced fat gain, it is evident that pregnancy may represent a risk factor for obesity (Scholl et al., 1995; Rooney et al., 2005; Amorim et al., 2007). Finally, our study was not intended to distinguish the effects of pregnancy from those of lactation. Lactation stimulates the secretion of prolactin and oxytocin (Augustine et al., 2008; Ladyman et al., 2010). In mice, lactation is a period of high metabolic demand due to the cost of milk production, which is compensated for by hyperphagia (Woodside et al., 2012; Zampieri et al., 2015; Teixeira et al., 2019a). Thus, lactation may have contributed to the growth of the females and should be considered alongside gestation when interpreting our results. Moreover, female mice housed with sterile males exhibit increased body weight and reduced lifespan (Garratt et al., 2020). Therefore, mating per se, even in the absence of fertilization, is sufficient to influence body weight and longevity, representing an additional factor—beyond pregnancy and lactation—that may have contributed to the parameters analyzed.

In conclusion, reproductive experience, including mating, pregnancy, and lactation, activates redundant, multi-hormonal anabolic programs involving growth factors and placental lactogens, among others. Our findings indicate that this coordinated endocrine-metabolic environment is sufficient to promote longitudinal skeletal growth, even in the absence of pituitary GH, in species with open growth plates. We also provided evidence that post-pregnancy growth can occur in women with severe IGHD. Thus, reproductive experience has many effects on the maternal organism.

Despite the lack of longitudinal growth in normal women, their bodies undergo significant and potentially permanent changes that need further study to understand the long-term impact on women’s health.

Material and Methods

Animals

In the current study, the following mouse strains were used: C57BL/6J (The Jackson Laboratory, Bar Harbor, ME; RRID:IMSR_JAX:000664), Ghrhrlit/lit(The Jackson Laboratory; RRID:IMSR_JAX:000533), AlbCre (The Jackson Laboratory; RRID:IMSR_JAX:003574), Ghrflox (List et al., 2014), ERαflox (The Jackson Laboratory; RRID:IMSR_JAX:032173), Nestincre (The Jackson Laboratory; RRID:IMSR_JAX:003771), and GHSR-deficient mice (Zigman et al., 2005). The generation, genotyping, and validation of the experimental mice were carried out as described in previous studies from our group (Frazao et al., 2013; Furigo et al., 2019; Teixeira et al., 2019a; Wasinski et al., 2020; Wasinski et al., 2022; Wasinski et al., 2023; de Sousa et al., 2025a; de Sousa et al., 2025b). Mice had ad libitum access to regular rodent chow (2.99 kcal/g, 9.4% kcal from fat, 236 g protein/kg; Nuvilab CR-1, Quimtia, Brazil) and filtered water. The experimental procedures were approved by the Ethics Committee on the Use of Animals of the Institute of Biomedical Sciences at the University of São Paulo.

Evaluation of pregnancy-induced growth in mice

The body composition (lean and fat mass) of adult (∼10-week-old) female mice was determined by time-domain nuclear magnetic resonance using the LF50 body composition mice analyzer (Bruker, Germany). Subsequently, the females were mated with sexually experienced males. Upon detection of pregnancy, females were placed in individual cages and maintained with their litters for 3-4 weeks. After weaning, the dams were regrouped in cages with up to five females. Body composition was reassessed 2 weeks after the last contact with the pups (∼20-week-old), which we consider the lactation recovery period. At the same time, body composition was analyzed in age-matched virgin females as a control group. In some experiments, the primiparous females were allowed to mate again, and the entire procedure was repeated to determine the effects of the second pregnancy and lactation cycle. After the last body composition assessment, WT and Ghrhrlit/lit mice were euthanized in a CO2 chamber, and naso-anal length was measured, as were the masses of the liver, heart, kidneys, and brain. Specifically, for GHSR KO mice and their respective controls, lean mass was calculated as total body weight minus fat mass (the sum of periuterine, periovarian, retroperitoneal, inguinal, posterior subcutaneous, and anterior subcutaneous fat depots).

Evaluation of pulsatile GH secretion and IGF-1 levels

To adapt the mice to the tail-tip blood sampling procedure, females were acclimated daily for 2-3 weeks before being placed in breeding. This adaptation continued while the females were housed with males. In pregnant females (gestational age 14-17 days) and virgin controls, a total of 36 blood samples were taken from the tail tip every 10 minutes. The procedure initially involved clipping a small portion (1 mm) from the tail tip with a surgical blade. Each blood sample (5 μL for WT and 2 μL for Ghrhrlit/lit) was then transferred to a tube containing PBS with 0.05% Tween-20. After each draw, gentle fingertip pressure was applied to the tail to halt bleeding. The mice continued to move freely within their home cages, with unlimited access to food and water. Immediately after collection, samples were placed on dry ice and stored at -80°C.

An enzyme-linked immunosorbent assay (ELISA) was performed to measure blood GH levels, using a 1:22 final dilution, and the following antisera: monkey anti-rat GH as the capture antibody (1:50,000; NHPP, Cat# AFP411S; RRID:AB_2665564) and rabbit anti-rat GH as the detection antibody (1:100,000; NHPP, Cat# AFP5672099, RRID:AB_2721132). GH pulses were detected using the DynPeak pulse detection algorithm’s default settings (Vidal et al., 2012). More details about this ELISA and the analysis of the GH secretion pattern are found in previous publications (Wasinski et al., 2020; Wasinski et al., 2022; de Sousa et al., 2024; de Sousa et al., 2025b). Serum IGF-1 levels were measured with a commercial ELISA kit (#MG100; RRID:AB_2827989; R&D Systems, Minneapolis, MN, USA).

Western Blot

Liver proteins were extracted using RIPA buffer (Sigma-Aldrich, Cat# R0278), containing phosphatase (Sigma-Aldrich, Cat# P5726 and P0044) and protease (Sigma-Aldrich, Cat# P8340) inhibitors. Samples were fractionated by polyacrylamide gel electrophoresis containing SDS (SDS-PAGE) using a running buffer (25 mM Tris; 192 mM Glycine; 0.1% SDS). Then, proteins were transferred to a nitrocellulose membrane overnight at 15 V using transfer buffer (25 mM Tris; 192 mM Glycine; 20% Methanol).

Membranes were blocked at room temperature for 2 h in 5% skim milk diluted in TBS-T buffer (10 mM Tris–HCl; 150 mM NaCl; 0.05% Tween-20). Membranes were incubated overnight in primary antibodies diluted in TBS-T containing 3% albumin. Membranes were washed three times for 10 min with TBS-T and incubated at room temperature for 1 h with respective secondary antibodies, diluted in 5% skim milk in TBS-T. After three 10-minute washes with TBS-T, the membranes were treated with chemiluminescent reagents (Clarity Western ECL Substrate, Bio-Rad, Hercules, CA; Cat# 170-5060). Bands on the membranes were revealed by automatic imaging systems. The intensities of the bands were quantified by densitometry using ImageJ software (https://imagej.net/ij/) and normalized to total protein expression in the gel (Ponceau staining). Primary antibodies used for blotting were anti-phosphoTyr694-STAT5 antibody (Cell Signaling Technology, Beverly, MA; Cat# 9351; RRID:AB_2315225; 1:1,000), anti-STAT5b antibody (Santa Cruz, Dallas, TX; Cat# sc-377069; 1:1,000), and anti-β-actin (Santa Cruz; Cat# sc-47778, RRID:AB_626632; 1:2,500).

Gene expression

Total RNA from the liver was extracted with TRIzol (Invitrogen, Carlsbad, CA), followed by incubation in DNase I RNase-free (MilliporeSigma, St. Louis, MO, USA). Reverse transcription was performed using 2 µg of total RNA, SuperScript II Reverse Transcriptase (Invitrogen), and random primers p(dN)6 (MilliporeSigma). Real-time PCR was performed using the 7500TM Real-Time PCR System (Applied Biosystems, Warrington, UK) and SYBR Green Gene Expression PCR Master Mix (Applied Biosystems. The following primers were used: Actb (forward: gctccggcatgtgcaaag; reverse: catcacaccctggtgccta), Cish (forward: tcgggaatctgggtggtact; reverse: taggaatgtaccctccggca), Esr1 (forward: gcagatagggagctggttca; reverse: tggagattcaagtccccaaa), Ghr (forward: atcaatccaagcctggggac; reverse: acagctgaatagatcctgggg), Igf1 (forward: ccacactgacatgcccaaga; reverse: gtacttcctttccttctcctttgc), Prlr (forward: cagtaaatgccaccaacgaa; reverse: gaggaggctctggttcaaca), Ppia (forward: tatctgcactgccaagactgagt; reverse: cttcttgctggtcttgccattcc), Socs2 (forward: cgcgagctcagtcaaacag; reverse: aagttccttctggagcctctt), and Stat5b (forward: ggactccgtccttgataccg; reverse: tccatcgtgtcttccagatcg). Data were normalized to the Actb (β-actin) and Ppia expression. Relative quantification of mRNA was calculated by 2-ΔΔCt.

Statistical analysis

When comparing two groups, the data were analyzed using a two-tailed, unpaired Student’s t-test. When comparing three groups simultaneously, one-way ANOVA followed by the Newman-Keuls multiple comparisons test was used. Two-way ANOVA and Holm–Sidak’s multiple comparisons test were used to analyze two variables (pregnancy and mutation effects). Statistical analyses were conducted with Prism software (GraphPad, San Diego, CA). All results are presented as mean ± standard error of the mean.

eLife Assessment

This important study reports a novel phenomenon of maternal growth during pregnancy that is independent of growth hormone (GH), adding a new dimension to maternal biology of reproduction. The evidence is convincing and supported by state-of-the-art methodologies conducted in mice and persuasive observations in humans with hereditary isolated GH deficiency. Revised discussion should focus on possible mechanisms, including the role of IGF2, and on directions for future research.

Reviewer #1 (Public review):

This work evaluates the impact of reproductive history on growth, body weight and body composition in mammals. In mice, somatic growth is stimulated by the first pregnancy while the second pregnancy increases body weight mainly by increasing adiposity. To probe the role of pituitary growth hormone (GH), the key regulator of somatic growth in these processes, was addressed by comparing the impact of reproduction on growth in normal ("wild type") and genetically GH-deficient females and by detailed characterization of the profile of fluctuations in circulating GH levels in both types of animals. Additional studies addressed the possible role of other endocrine pathways (ghrelin and estrogen) in the pregnancy-related growth. Surprisingly, reproduction-related growth was independent of GH, ghrelin and estrogen. To determine whether these results may apply ("translate") to human physiology, data on various parameters of somatic growth were collected from women with hereditary GH deficiency. The findings indicate that GH-independent stimulation of growth by reproductive events also occurs in women.

Use of multiple animal models, rigorous characterization of GH levels in normal and GH-deficient females, and inclusion of data derived from a unique and well-characterised population of people with hereditary isolated GH deficiency and no GH replacement therapy are important strengths of these elegant and innovative studies. The results address a broader and clinically significant issue of permanent changes in body size, composition and function that result from pregnancy and lactation. This work also provides important background for further studies aimed at the identification of the mechanism involved and the role of specific reproductive events in the regulation of growth.

Reviewer #2 (Public review):

This manuscript describes the fascinating phenomenon of growth hormone (GH)-independent growth occurring in the mother during pregnancy. This growth was most pronounced in dwarf mice that are lacking the receptor for growth hormone-releasing hormone (GHRH) and therefore showing isolated GH deficiency. However, the pregnancy-induced growth could also be observed in wild-type mice, suggesting that it is a normal part of the maternal adaptation to pregnancy. The study falls short of identifying the mechanism(s) driving this pregnancy-induced growth response, but it certainly reveals a novel insight into maternal physiology. The authors have completed a range of experiments in mice to prove that, as well as being GH independent, the pregnancy-induced growth also did not require GH signaling in the liver (i.e. not another pregnancy-specific ligand operating through the GHR to promote IGF). They also provided complementary data from a population of humans with untreated isolated GH deficiency that are broadly consistent with the hypothesis. While it is important to consider the significant species differences between rodents and humans, both in terms of growth physiology and also in terms of evolution of placental somato-mammotrophic hormones, this unique population are a valuable resource and adds credence to the study. Overall, I find this a compelling research story, but disappointingly unfinished. There are some areas where additional information could improve the ability to interpret the data, and some additional concepts that could be considered in the discussion. There are also areas where additional experiments might provide important insights. However, I think that such suggestions can be considered as appropriate for future research, rather than delaying consideration of the current manuscript.

Main comments:

(1) Data in Figure 1 are remarkable - not so much the growth in pregnancy in the wildtype mice, because while elevated GH is well known in pregnancy, but growth in the dwarf mice is indicative of GH-independent growth. From these data, it seems that there is good evidence that growth in pregnancy is an adaptive function. However, it is possible that growth is achieved in dwarf mice and that in wildtype mice may have been mediated through different mechanisms. The dwarf mice showed an increase in liver and plasma IGF1, suggestive of an additional ligand driving IGF in pregnancy. One could hypothesize that such an effect could be mediated by an additional pregnancy-specific ligand activating the GH receptor. In humans, placental growth hormone could be such a ligand, but as far as we know, there is no placental GH in mice. In contrast, the wildtype animals showed suppression of liver and circulating IGF1, and low levels of pSTAT5 in the liver during pregnancy. These data (in Figure 5) are very surprising. Given the high circulating GH in pregnancy, as well as high placental lactogen (which would be expected to activate STAT5 in the liver through the Prlr), the low levels of pSTAT5 are unexpected and would seem to indicate some sort of acquired insensitivity to GH. Is this entirely driven by down-regulation of STAT5b protein, or could there be activation of other, negative regulators of STAT signalling, such as SOCS? What is causing such a profound suppression of STAT5? Regardless of the mechanism, this suggests that pregnancy-induced growth in wildtype mice is independent of circulating IGF1 (potentially a different mechanism or in addition to that seen in IGHD mice).

The data shown in Figure 6 are a major strength of the study, showing that the pregnancy-induced changes are not specific to one particular transgenic model, but still occur in a variety of models affecting GH through different approaches. Given the pregnancy-specific nature of the changes, however, it seems an oversight not to have evaluated the role of placental lactogens. Prlr is highly expressed in the liver, but the function of this hormone in the liver is not well established. Could the extremely high levels of PL be mediating this growth response? Given the low expression of STAT5 in the liver and the fact that plasma IGF1 is not markedly elevated, it seems more likely that this growth response may be mediated by locally produced IGF1 in target tissues.

I think these possibilities could be addressed by an expanded discussion of species variation in placental hormones, to highlight that humans have expansion of the GH locus, but rodents have expansion of the prolactin axis (see Soares, M. J. The prolactin and growth hormone families: pregnancy-specific hormones/cytokines at the maternal-fetal interface. Reprod Biol Endocrinol 2, 51, 2004). Importantly, placental GH and chorionic somatomammotropins (CSM) in humans are all variants of the GH gene, but CSM have preferential activity at Prlr. This seems to be a fundamental species difference in pregnancy biology, but has been interpreted as an example of convergent evolution, with conservation of prolactin and GH-like functions at the maternal-fetal interface, mediated by different mechanisms, likely contributing to the metabolic adaptations of the mother (see Newbern D, Freemark M. Placental hormones and the control of maternal metabolism and fetal growth. Curr Opin Endocrinol Diabetes Obes. 2011; 18: 409-416). While the preceding function has focused on explaining the evolution of placental lactogens (either prolactin or GH variants), the present data suggest that there are also mechanisms to maintain growth in pregnancy, independent of GH (even in the absence of a placental GH).

(2) The human data are very interesting, and my initial impression was that it seemed unlikely to be the same phenomenon. Was there any real evidence for "growth" in pregnancy? Pubertal maturation of long bone growth might be expected to prevent further growth in adulthood. However, these issues were appropriately discussed, and it seems well justified to evaluate this unique population of women with IGHD who underwent pregnancy. It would be very interesting to know if these women experienced elevated IGF1 during pregnancy, indicative of placental GH contributing to growth. Mechanistically, this might be more like the dwarf mouse situation of IGHD, that the situation in wildtype mice (associated with liver insensitivity to GH and low IGF1).

(3) It would be useful to include investigations that isolate the effects of pregnancy and the placental hormones. Such studies could include evaluating growth in pseudopregnant mice with IGHD (pregnancy-like changes in hormones but lacking the placental contribution) and in IGHD animals that experience pregnancy but not lactation (pups removed at birth). I accept that this might be too large an additional study to add for the present manuscript.

(4) It is an important and translationally relevant observation that pregnancy increased the risk of long-term weight gain, and that after the first pregnancy, the pregnancy-induced growth response was more directed to promoting fat deposition. Does this provide any mechanistic insight? Could a metabolic adaptation result in growth?

Reviewer #3 (Public review):

Summary:

The study describes an increase in body growth and body composition in both mice and women. In mice, the impact on growth is mainly seen during the first pregnancy, and the changes postpartum on body composition are also different during the first and second pregnancies. The study has used various knock-out models in the growth hormone axis to understand these changes as well as some gene expression analysis related to GH, IGF-1 and estrogen signalling pathways.

Strengths:

(1) The inclusion of various knock-out mouse models that allow for exploration of mechanisms related to the above-mentioned changes.

(2) The investigation of gene expression of GHR, IGF-1R and ER pathways.

Weaknesses:

The human findings are dependent on the patient's recollection of bodily changes after their pregnancies.

Conclusion:

The authors have partly achieved their aim of describing changes in growth and body composition that remain after pregnancy and the mechanisms behind these changes. This study may have importance for a wide variety of research areas as well as in the clinical setting. The study is also unique in its attempt to bridge findings in mice to a unique human model of congenital GH deficiency.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment:

This important study reports a novel phenomenon of maternal growth during pregnancy that is independent of growth hormone (GH), adding a new dimension to maternal biology of reproduction. The evidence is convincing and supported by state-of-the-art methodologies conducted in mice and persuasive observations in humans with hereditary isolated GH deficiency. Revised discussion should focus on possible mechanisms, including the role of IGF2, and on directions for future research.

We thank the editors for these valuable suggestions. In response, we assessed hepatic Igf2 mRNA expression. As shown in Author response image 1, hepatic Igf2 mRNA expression was higher in Ghrhrlit/lit mice but was not significantly affected by pregnancy. However, we recognize that IGF-2 is produced by multiple tissues, particularly at the maternal-fetal interface, as well as by other maternal tissues, exerting endocrine or paracrine effects. Therefore, we believe that hepatic Igf2 expression alone provides only a limited view of the potential contribution of IGF-2 to the maternal adaptations induced by pregnancy. For this reason, we chose not to include these data in the revised manuscript. Instead, we expanded the Discussion to address the potential role of IGF-2 in the observed phenotypes and to highlight important directions for future research, as suggested.

Public Reviews:

Reviewer #1 (Public review):

This work evaluates the impact of reproductive history on growth, body weight and body composition in mammals. In mice, somatic growth is stimulated by the first pregnancy while the second pregnancy increases body weight mainly by increasing adiposity. To probe the role of pituitary growth hormone (GH), the key regulator of somatic growth in these processes, was addressed by comparing the impact of reproduction on growth in normal ("wild type") and genetically GH-deficient females and by detailed characterization of the profile of fluctuations in circulating GH levels in both types of animals. Additional studies addressed the possible role of other endocrine pathways (ghrelin and estrogen) in the pregnancy-related growth. Surprisingly, reproduction-related growth was independent of GH, ghrelin and estrogen. To determine whether these results may apply ("translate") to human physiology, data on various parameters of somatic growth were collected from women with hereditary GH deficiency. The findings indicate that GH-independent stimulation of growth by reproductive events also occurs in women.

Use of multiple animal models, rigorous characterization of GH levels in normal and GH-deficient females, and inclusion of data derived from a unique and well-characterised population of people with hereditary isolated GH deficiency and no GH replacement therapy are important strengths of these elegant and innovative studies. The results address a broader and clinically significant issue of permanent changes in body size, composition and function that result from pregnancy and lactation. This work also provides important background for further studies aimed at the identification of the mechanism involved and the role of specific reproductive events in the regulation of growth.

We sincerely thank the reviewer for this thoughtful and highly positive evaluation of our work and for recognizing its novelty, rigor, and potential physiological and clinical significance.

Reviewer #2 (Public review):

This manuscript describes the fascinating phenomenon of growth hormone (GH)-independent growth occurring in the mother during pregnancy. This growth was most pronounced in dwarf mice that are lacking the receptor for growth hormone-releasing hormone (GHRH) and therefore showing isolated GH deficiency. However, the pregnancy-induced growth could also be observed in wild-type mice, suggesting that it is a normal part of the maternal adaptation to pregnancy. The study falls short of identifying the mechanism(s) driving this pregnancy-induced growth response, but it certainly reveals a novel insight into maternal physiology. The authors have completed a range of experiments in mice to prove that, as well as being GH independent, the pregnancy-induced growth also did not require GH signaling in the liver (i.e. not another pregnancy-specific ligand operating through the GHR to promote IGF). They also provided complementary data from a population of humans with untreated isolated GH deficiency that are broadly consistent with the hypothesis. While it is important to consider the significant species differences between rodents and humans, both in terms of growth physiology and also in terms of evolution of placental somato-mammotrophic hormones, this unique population are a valuable resource and adds credence to the study. Overall, I find this a compelling research story, but disappointingly unfinished. There are some areas where additional information could improve the ability to interpret the data, and some additional concepts that could be considered in the discussion. There are also areas where additional experiments might provide important insights. However, I think that such suggestions can be considered as appropriate for future research, rather than delaying consideration of the current manuscript.

Main comments:

(1) Data in Figure 1 are remarkable - not so much the growth in pregnancy in the wildtype mice, because while elevated GH is well known in pregnancy, but growth in the dwarf mice is indicative of GH-independent growth. From these data, it seems that there is good evidence that growth in pregnancy is an adaptive function. However, it is possible that growth is achieved in dwarf mice and that in wildtype mice may have been mediated through different mechanisms. The dwarf mice showed an increase in liver and plasma IGF1, suggestive of an additional ligand driving IGF in pregnancy. One could hypothesize that such an effect could be mediated by an additional pregnancy-specific ligand activating the GH receptor. In humans, placental growth hormone could be such a ligand, but as far as we know, there is no placental GH in mice. In contrast, the wildtype animals showed suppression of liver and circulating IGF1, and low levels of pSTAT5 in the liver during pregnancy. These data (in Figure 5) are very surprising. Given the high circulating GH in pregnancy, as well as high placental lactogen (which would be expected to activate STAT5 in the liver through the Prlr), the low levels of pSTAT5 are unexpected and would seem to indicate some sort of acquired insensitivity to GH. Is this entirely driven by down-regulation of STAT5b protein, or could there be activation of other, negative regulators of STAT signalling, such as SOCS? What is causing such a profound suppression of STAT5? Regardless of the mechanism, this suggests that pregnancy-induced growth in wildtype mice is independent of circulating IGF1 (potentially a different mechanism or in addition to that seen in IGHD mice).

We thank the reviewer for this insightful comment. The mechanisms underlying the reduced hepatic STAT5b protein expression observed during pregnancy remain unknown. Previous studies have also reported desensitization of hepatic JAK2/STAT5 GH signaling during pregnancy, associated with increased expression of negative regulators of STAT signaling, including CIS (Miquet et al., Am J Physiol Endocrinol Metab 289–E607, 2005). In the current manuscript, we also observed increased Cish mRNA expression in the liver of pregnant mice. We have incorporated this information into the revised manuscript. In addition, we expanded the Discussion to acknowledge the possibility that pregnancy-induced growth may occur through distinct mechanisms in wild-type and GH-deficient mice and that, in wild-type mice, this response may involve liver-derived IGF-1-independent mechanisms.

The data shown in Figure 6 are a major strength of the study, showing that the pregnancy-induced changes are not specific to one particular transgenic model, but still occur in a variety of models affecting GH through different approaches. Given the pregnancy-specific nature of the changes, however, it seems an oversight not to have evaluated the role of placental lactogens. Prlr is highly expressed in the liver, but the function of this hormone in the liver is not well established. Could the extremely high levels of PL be mediating this growth response? Given the low expression of STAT5 in the liver and the fact that plasma IGF1 is not markedly elevated, it seems more likely that this growth response may be mediated by locally produced IGF1 in target tissues.

We thank the reviewer for this important suggestion. Although we were not able to investigate the role of prolactin receptor signaling using mouse models in the present study, we agree that prolactin may contribute to pregnancy-induced growth. Accordingly, we revised the Discussion to acknowledge this possibility and to include the potential roles of placental lactogens and locally produced IGF-1 as additional mechanisms that may underlie pregnancy-induced body growth.

I think these possibilities could be addressed by an expanded discussion of species variation in placental hormones, to highlight that humans have expansion of the GH locus, but rodents have expansion of the prolactin axis (see Soares, M. J. The prolactin and growth hormone families: pregnancy-specific hormones/cytokines at the maternal-fetal interface. Reprod Biol Endocrinol 2, 51, 2004). Importantly, placental GH and chorionic somatomammotropins (CSM) in humans are all variants of the GH gene, but CSM have preferential activity at Prlr. This seems to be a fundamental species difference in pregnancy biology, but has been interpreted as an example of convergent evolution, with conservation of prolactin and GH-like functions at the maternal-fetal interface, mediated by different mechanisms, likely contributing to the metabolic adaptations of the mother (see Newbern D, Freemark M. Placental hormones and the control of maternal metabolism and fetal growth. Curr Opin Endocrinol Diabetes Obes. 2011; 18: 409-416). While the preceding function has focused on explaining the evolution of placental lactogens (either prolactin or GH variants), the present data suggest that there are also mechanisms to maintain growth in pregnancy, independent of GH (even in the absence of a placental GH).

Thank you very much for this observation and suggestion. In the Introduction, we succinctly mention species-specific differences in the GH locus (see Introduction). Since our study focused primarily on mice, we believe expanding on this information would detract from the study's focus, especially given that such details are available in other publications. However, we have added a sentence to the discussion noting that, despite significant species-specific differences in the GH locus, there has been evolutionary conservation in the maintenance of pregnancy-induced growth, potentially involving distinct mechanisms in rodents and humans, though both appear to be GH-independent.

“However, important differences are observed between humans and rodents. Humans, but not rodents, possess the GH2 gene, which encodes the placental growth hormone (GH) variant, whose production progressively replaces pituitary GH (derived from the GH1 gene) secretion during pregnancy (Liao et al., 2018). In contrast, rodents maintain pituitary GH secretion throughout gestation (Gatford et al., 2017; Liao et al., 2018). Additionally, human GH, both pituitary and placental variants, as well as chorionic somatomammotropins, can activate both the GH receptor (GHR) and the prolactin receptor (PrlR), explaining their lactogenic effects. In contrast, murine GH activates only the GHR (Bartke and Kopchick, 2015).”

(2) The human data are very interesting, and my initial impression was that it seemed unlikely to be the same phenomenon. Was there any real evidence for "growth" in pregnancy? Pubertal maturation of long bone growth might be expected to prevent further growth in adulthood. However, these issues were appropriately discussed, and it seems well justified to evaluate this unique population of women with IGHD who underwent pregnancy. It would be very interesting to know if these women experienced elevated IGF1 during pregnancy, indicative of placental GH contributing to growth. Mechanistically, this might be more like the dwarf mouse situation of IGHD, that the situation in wildtype mice (associated with liver insensitivity to GH and low IGF1).

We thank the reviewer for this comment. Unfortunately, the pregnancies of the women with IGHD occurred with limited medical assistance, and no additional clinical information is available, particularly regarding changes in circulating IGF-1 levels during pregnancy. Therefore, the evidence presented and discussed in the current manuscript represents the only information currently available on pregnancy-induced growth in this unique population.

(3) It would be useful to include investigations that isolate the effects of pregnancy and the placental hormones. Such studies could include evaluating growth in pseudopregnant mice with IGHD (pregnancy-like changes in hormones but lacking the placental contribution) and in IGHD animals that experience pregnancy but not lactation (pups removed at birth). I accept that this might be too large an additional study to add for the present manuscript.

We thank the reviewer for this valuable suggestion and agree that this experiment could provide valuable information regarding the role of placental hormones in pregnancy-induced growth. Interestingly, previous work has shown that female mice housed with vasectomized males exhibit increased body weight compared with non-reproductive females, although this increase is substantially smaller than that observed in females with successive litters (Garratt et al., Proc Natl Acad Sci U S A 117:15748–15754, 2020). We discuss this study in the current manuscript. However, this study did not provide additional information on whether this increase reflects greater adiposity, enhanced body growth, or both. Unfortunately, we are unable to perform these time-intensive experiments. We appreciate the reviewer's understanding and agree that this important question should be addressed in future studies.

(4) It is an important and translationally relevant observation that pregnancy increased the risk of long-term weight gain, and that after the first pregnancy, the pregnancy-induced growth response was more directed to promoting fat deposition. Does this provide any mechanistic insight? Could a metabolic adaptation result in growth?

We thank the reviewer for this thoughtful comment. At present, we do not have additional data to explain the mechanisms underlying the pregnancy-induced increases in body growth and adiposity or to determine whether these processes are mechanistically linked. In the revised manuscript, we have expanded the Discussion to address potential mechanisms that may contribute to the observed post-pregnancy growth, including hormonal and metabolic factors, while acknowledging that these possibilities remain speculative and require further investigation.

Reviewer #3 (Public review):

Summary:

The study describes an increase in body growth and body composition in both mice and women. In mice, the impact on growth is mainly seen during the first pregnancy, and the changes postpartum on body composition are also different during the first and second pregnancies. The study has used various knock-out models in the growth hormone axis to understand these changes as well as some gene expression analysis related to GH, IGF-1 and estrogen signalling pathways.

Strengths:

(1) The inclusion of various knock-out mouse models that allow for exploration of mechanisms related to the above-mentioned changes.

(2) The investigation of gene expression of GHR, IGF-1R and ER pathways.

Weaknesses:

The human findings are dependent on the patient's recollection of bodily changes after their pregnancies.

Conclusion:

The authors have partly achieved their aim of describing changes in growth and body composition that remain after pregnancy and the mechanisms behind these changes. This study may have importance for a wide variety of research areas as well as in the clinical setting. The study is also unique in its attempt to bridge findings in mice to a unique human model of congenital GH deficiency.

We sincerely thank the reviewer for the thoughtful evaluation of our study, for recognizing its novelty and potential significance, and for highlighting the strengths of our experimental approach. We have revised the manuscript to further acknowledge the limitations of the human data, as suggested.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

The findings presented in this study are novel, exciting and important. I believe it would help the readers to include a diagram showing the timing of the various measurements and tissue collection in relation to conception, parturition and weaning. It would also be helpful to expand the discussion to include the author's thoughts on the possible role of various placental hormones, pituitary prolactin, metabolic changes and/or other factors in the observed growth. Finally, humans from whom the data were obtained should probably be referred to as individuals or subjects rather than patients.

We thank the reviewer for these helpful suggestions. We included Figure 7 in the revised manuscript, which summarizes the study's experimental timeline. Additionally, we expanded the Discussion as recommended and replaced the term "patients" with "subjects" throughout the manuscript.

Reviewer #2 (Recommendations for the authors):

In addition to my comment above about addressing the different evolution of placental hormones, the following minor suggestions are provided:

(1) Abstract line 23 "most" could be "some" (because lots of evidence that pregnancy-induced adaptations are long-term - e.g. brain structure, maternal behavior, endocrine changes) - see below.

Done.

(2) Is "reproductive experience" the best term? Or is it "pregnancy"? (likely specifically due to the presence of placental hormones and/or maternal patterns of hormones, rather than other aspects of "reproductive experience) e.g. mating, parturition or lactation?

We appreciate the reviewer's point. However, because we cannot exclude potential contributions of mating, parturition, or lactation to the observed increases in body growth and adiposity, we prefer to retain the term "reproductive experience," as the mice in the present study underwent all of these processes rather than pregnancy alone.

(3) Potentially, the introduction could add references to brain changes in humans in pregnancy (this is a particularly topical area of research where long-term changes have been reported; e.g. Pritschet L, Taylor CM, Cossio D, Faskowitz J, Santander T, Handwerker DA, Grotzinger H, Layher E, Chrastil ER, Jacobs EG. Neuroanatomical changes observed over the course of a human pregnancy. Nat Neurosci. 2024 Nov;27(11):2253-2260; Martínez-García M, Paternina-Die M, Barba-Müller E, Martín de Blas D, Beumala L, Cortizo R, Pozzobon C, Marcos-Vidal L, Fernández-Pena A, Picado M, Belmonte-Padilla E, Massó-Rodriguez A, Ballesteros A, Desco M, Vilarroya Ó, Hoekzema E, Carmona S. Do Pregnancy-Induced Brain Changes Reverse? The Brain of a Mother Six Years after Parturition. Brain Sci. 2021 Jan 28;11(2):168. There is also a significant literature on long-term effects of reproductive experience (Bridges RS. Long-term alterations in neural and endocrine processes induced by motherhood in mammals. Horm Behav. 2016 Jan;77:193-203).

We thank the reviewer for this helpful suggestion. Accordingly, we incorporated these references into the Introduction to further emphasize the long-term effects of reproductive experience.

(4) Figure 1: Why show body weight as a "delta" for primiparous animals, and "absolute "values for multiparous? Similarly, it seems that body length is an informative readout for these experiments, but this is only reported for one model, and most models are only shown as body weight and lean mass.

We thank the reviewer for this comment. For the primiparous mice, we chose to present the change (Δ) in body weight and lean mass because it more directly illustrates the effects of the first reproductive experience relative to each animal's baseline while avoiding redundant presentation of both absolute and Δ values, as both analyses yielded the same statistical conclusions. For the multiparous mice, absolute values were presented to facilitate comparison with the corresponding images and to allow direct comparison of body size between wild-type and Ghrhrlit/lit mice. Regarding body length, these measurements were collected only in the multiparous cohort and were not obtained after the first reproductive experience. Therefore, we are unable to provide body length data for the primiparous groups.

(5) It seems it would be better to present all of the mouse data first, then the supporting human data.

Done.

(6) The manuscript includes a high-quality investigation of GH secretion - particularly the pulsatile secretion of GH. Is it possible to have matching IGF data (at least multiple time points in pregnancy)?

We agree with the reviewer that serial measurements of circulating IGF-1 during pregnancy would provide valuable additional information. Unfortunately, this analysis is not feasible because the limited blood volume collected during the study did not allow sufficient sample storage for additional measurements.

(7) Line 221 - "GHRH signaling is required for pregnancy-induced increases in GH secretion...". Can you actually make that conclusion, or is it possible that the lack of GHRH developmentally has permanently altered the pituitary gland, so it cannot respond to other GH secretagogues?

We thank the reviewer for this important comment. We agree that the congenital absence of GHRH signaling may cause developmental alterations in the pituitary that impair the response to other GH secretagogues. Therefore, we cannot distinguish whether the lack of pregnancy-induced increases in GH secretion reflects the absence of GHRH signaling per se or the developmental defects resulting from GHRH deficiency. We removed the aforementioned phrase from Results and have revised the Discussion to acknowledge this important limitation and to clarify that the observed phenotype may be secondary to impaired pituitary development.

(8) Line 238 - "GHR signaling stimulates hepatic insulin-like growth factor 1 (IGF-1) secretion, particularly by activating the signal transducer and activator of transcription 5b (STAT5b) pathway" for this point, should probably cite: Davey HW, Xie T, McLachlan MJ, Wilkins RJ, Waxman DJ, Grattan DR. STAT5b is required for GH-induced liver IGF-I gene expression. Endocrinology. 2001 Sep;142(9):3836-41.

Thank you for this suggestion; we added this reference.

(9) Line 240 - "GH action in the liver is not only controlled by the pattern of GH secretion...". should probably cite Waxman's work regarding patterns of GH secretion (e.g reviewed in Waxman DJ, O'Connor C. Growth hormone regulation of sex-dependent liver gene expression. Mol Endocrinol. 2006 Nov;20(11):2613-29.).

This reference was included in this sentence.

(10) Methods: Check statistics - some of the analysis should have been repeated measures?

We thank the reviewer for this comment. We carefully re-evaluated all statistical analyses and did not identify any experiment for which a repeated-measures analysis would have been appropriate. Therefore, no changes to the statistical analyses were necessary.

Reviewer #3 (Recommendations for the authors):

(1) Introduction: The context of the sentence "Furthermore, historically, females have been underrepresented in studies" is unclear.

We thank the reviewer for this observation. Our intention was to emphasize that, because pregnancy is unique to females and females have historically been underrepresented in biomedical research, pregnancy-related physiology has received comparatively less attention. However, we agree that this statement was not essential in this context. Thus, we have removed it from the revised manuscript.

(2) Introduction: Based on the paper, the main aim seems to be to study the lasting effects of pregnancy on the mother and not primarily to study the hormonal changes during pregnancy.

We thank the reviewer for this helpful observation. We agree that our study primarily focuses on the long-term effects of pregnancy on the maternal organism rather than the hormonal changes that occur during pregnancy. Accordingly, we revised this sentence to better reflect the main objective of the study: “Therefore, the goal of this study is to investigate the lasting effects of pregnancy on the maternal organism and the possible involvement of the somatotrophic axis in these changes.”

(3) Other endocrine factors could affect the changes seen after pregnancy in mice and humans, such as the insulin level and activation of the IR.

We thank the reviewer for this valuable suggestion. In the revised manuscript, we expanded the Discussion to acknowledge the potential contribution of several endocrine factors, including insulin signaling, to pregnancy-induced growth.

(4) The receptor activation in tissues other than the liver has not been studied, such as in the growth plate, adipose tissue and muscle.

We thank the reviewer for this important comment. We agree with this point and have addressed this possibility in the revised Discussion.

(5) Information on the age of the patient at the time of pregnancy could be included as well as the patient's height and weight before pregnancy and at follow-up.

This information was added as Table 1.

Version 2

Abstract

Pregnancy leads to many adaptations in the maternal body, some of which are reversible. However, reproductive experience can also result in permanent effects. Here, we investigated how pregnancy influences the somatotrophic system and the lasting effects of reproductive experience on the maternal organism. Reproductive experience induced a pronounced increase in lean body mass and longitudinal growth in both wild-type and growth hormone (GH)-deficient mice compared with age-matched virgins. Body growth was primarily observed during the first pregnancy, whereas a second gestation was mostly associated with increased adiposity. Increased GH secretion was observed in pregnant wild-type mice but not in pregnant GHRHR-deficient mice. Pregnancy-induced body growth is preserved despite disruption of GH-, ghrelin-, and estrogen-related signaling pathways. Data from a cohort of women with isolated GH deficiency (IGHD) caused by a loss-of-function mutation in the GHRHR gene revealed that nulliparous women were 7 cm shorter than those with one or more pregnancies. In conclusion, reproductive experience induces permanent changes in the maternal organism, promoting body growth in models that allow this response. Pregnancy-induced body growth appears to be independent of GH action. These findings underscore the need for further studies to investigate the long-lasting consequences of reproductive experience in females.

Introduction

Pregnancy is a unique state in which numerous physiological adaptations are necessary in the maternal organism to allow proper fetal growth and to prepare for the lactation period (Augustine et al., 2008). Among these adaptations, metabolic changes are particularly prominent, including alterations in appetite and glucose homeostasis, as well as increased adiposity (Augustine et al., 2008; Ladyman et al., 2010; Zampieri et al., 2015; Teixeira et al., 2019a). Pregnancy also affects neuroplasticity and causes behavioral and emotional alterations (Larsen and Grattan, 2012; Georgescu et al., 2021). Several of these changes are reversible. However, pregnancy may lead to long-term or permanent consequences. For example, prior pregnancy experience reduces the latency to display maternal behavior in rats (Bridges, 1978), indicating long-lasting behavioral adaptations strongly shaped by prior hormonal exposure. Primiparous mice show increased body weight and reduced ambulatory activity compared with age-matched virgin animals, even several months after last contact with the pups (Ladyman et al., 2018; Teixeira et al., 2019b; Ladyman et al., 2021). Body weight retention is frequently observed after pregnancy, especially for women who experienced excessive gestational weight gain (Scholl et al., 1995; Rooney et al., 2005). Thus, pregnancy may represent a risk factor for obesity in women (Amorim et al., 2007). Additionally, reproductive experience induces persistent structural remodeling of the human brain, including reductions in gray matter volume and cortical thickness that remain evident for at least six years after parturition (Martinez-Garcia et al., 2021; Pritschet et al., 2024). These findings indicate that motherhood is associated with long-term adaptations in both neural and endocrine function (Bridges, 2016).

Pregnancy-induced adaptations are thought to be largely driven by the marked rise in circulating hormones during gestation (Augustine et al., 2008), including progesterone and estrogens, as well as placentally derived analogues of pituitary hormones, such as chorionic gonadotropin and placental lactogens (also referred to as chorionic somatomammotropins). However, important differences are observed between humans and rodents. Humans, but not rodents, possess the GH2 gene, which encodes the placental growth hormone (GH) variant, whose production progressively replaces pituitary GH (derived from the GH1 gene) secretion during pregnancy (Liao et al., 2018). In contrast, rodents maintain pituitary GH secretion throughout gestation (Gatford et al., 2017; Liao et al., 2018). Additionally, human GH, both pituitary and placental variants, as well as chorionic somatomammotropins, can activate both the GH receptor (GHR) and the prolactin receptor (PrlR), explaining their lactogenic effects. In contrast, murine GH activates only the GHR (Bartke and Kopchick, 2015).

Despite species-specific differences, it is evident that hormonal exposure during pregnancy not only transiently shapes maternal physiology but may also exert lasting effects on health across the lifespan. However, the chronic consequences of reproductive experience remain poorly understood. Given that most women will experience at least one pregnancy during their lifetime, elucidating how reproductive experience influences long-term physiology and health is of substantial biological and clinical importance. Therefore, the goal of this study is to investigate the long-term effects of reproductive experience on the maternal organism and the potential involvement of the somatotrophic axis in these adaptations.

Results

Pregnancy promotes permanent body growth in wild-type and GH-deficient mice

To investigate the consequences of reproductive experience, body weight and composition of C57BL/6J wild-type (WT) mice were initially measured before mating. Pregnant mice gave birth in individual cages and were maintained with their litters for 3 weeks. Two weeks after weaning (the recovery period from lactation), the females were reevaluated to determine changes relative to baseline. The control group consisted of age-matched virgin females. These assessments were also carried out in Ghrhrlit/lit mice, which carry a null mutation in the gene encoding the GHRH receptor (GHRHR) and are therefore GH-deficient and dwarf. Reproductive experience caused a robust increase in body mass (Figure 1A) and lean mass (Figure 1B) in both control and Ghrhrlit/lit mice compared with virgin groups. No changes in fat mass were observed between groups (Figure 1C). A new evaluation was performed 2 weeks after the second pregnancy and lactation cycle. Body length measurements revealed a significant increase in multiparous WT and Ghrhrlit/lit mice compared with age-matched virgin animals (Figure 1D-E). In addition, multiparous mice showed higher body and lean mass than virgin animals (Figure 1F-G). Despite pregnancy-induced growth, Ghrhrlit/lit mice remained smaller than WT females (Figure 1D-G). Fat mass increased after 2 pregnancies in both WT and Ghrhrlit/litmice compared with their respective age-matched virgin groups (Figure 1H). Multiparous WT and Ghrhrlit/lit mice also had increased liver (Figure 1I) and kidney (data not shown) masses compared with virgin groups. Heart and brain masses increased only in multiparous Ghrhrlit/litmice compared with virgin Ghrhrlit/lit females, whereas these tissues did not change in WT mice (data not shown). Thus, pregnancy promotes body growth in WT and GH-deficient females.

Pregnancy promotes permanent body growth in wild-type and GH-deficient mice.

(A-C) Differences relative to baseline in body, lean, and fat mass in virgin WT (n = 15), primiparous wild-type (WT; n = 13), virgin Ghrhrlit/lit(n = 6), and primiparous Ghrhrlit/lit (n = 13). (D) Representative image showing differences in longitudinal growth between virgin and multiparous WT and Ghrhrlit/lit mice. (E-I) Body length, body weight, lean mass, fat mass, and liver mass in multiparous WT (n = 13) and Ghrhrlit/lit (n = 13) mice, relative to age-matched virgin WT (n = 15) and Ghrhrlit/lit (n = 12) mice. *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Growth is primarily induced during the first pregnancy

The following experiment tested whether a second pregnancy could also increase parameters indicative of body growth. Unlike the body weight and lean mass gains observed after the first pregnancy, WT mice subjected to a second pregnancy cycle showed no increases in body or lean mass compared with age-matched WT females (Figure 2A-B). Because females undergoing a second pregnancy are inherently older than during their first gestation, we assessed whether a first pregnancy occurring at a comparable age to the second would influence body weight. WT females that had a late first pregnancy (age-matched first pregnancy group) showed increased body and lean mass gain compared with age-matched virgin and second pregnancy groups (Figure 2A-B). Notably, the gains in body and lean mass induced by the first pregnancy were similar among the youngest and oldest females compared with their respective age-matched groups of virgin females (Figure 2A-B). No differences in fat mass gain were observed between the groups (data not shown). In contrast to WT mice, the second pregnancy resulted in a significant increase in body and lean mass in Ghrhrlit/lit females compared with age-matched virgin animals (Figure 2C-D). However, the magnitude of the increase in the second pregnancy was significantly lower than in the first (Figure 2C-D), highlighting that growth occurs primarily during the first pregnancy.

Growth is primarily induced during the first pregnancy.

(A-B) Changes in body weight and lean mass in wild-type (WT) virgins (n = 15), after the first pregnancy (n = 13), in virgins matched for age at the second pregnancy (n = 15), after the second pregnancy (n = 13), and after a late first pregnancy (n = 8). (C-D) Changes in body weight and lean mass in Ghrhrlit/litvirgins (n = 6), after the first pregnancy (n = 13), in virgins matched for age at the second pregnancy (n = 12), and after the second pregnancy (n = 13). *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using unpaired two-tailed Student’s t-test or one-way ANOVA and Newman-Keuls multiple comparisons test.

Increased GH secretion in pregnant WT mice but not in Ghrhrlit/litfemales

Because GH is the major circulating factor that induces longitudinal growth in mammals (Zhou et al., 1997; Dehkhoda et al., 2018), we investigated whether WT and Ghrhrlit/lit mice exhibit alterations in GH secretion during pregnancy (gestational age 14-17 days). Confirming previous studies (Gatford et al., 2017; Kaur et al., 2020; Wasinski et al., 2022), pregnancy increased GH secretion in WT mice compared with virgin animals (Figure 3A, C). This increase was associated with higher GH pulse frequency and elevations in both pulsatile and basal secretion (Figure 3C-G). The contribution of basal to total GH secretion increased in pregnant WT mice (Figure 3H). As expected, non-pregnant Ghrhrlit/lit mice showed blunted GH secretion (Figure 3B). Importantly, pregnancy did not stimulate GH secretion in Ghrhrlit/lit mice (Figure 3C-H).

Increased GH secretion in pregnant WT mice but not in pregnant Ghrhrlit/lit mice.

(A-B) Representative examples of the pattern of GH secretion in two pregnant and virgin WT mice and in two pregnant and virgin Ghrhrlit/lit mice. (C-H). Total GH secretion, GH pulse frequency, pulsatile GH secretion, GH pulse amplitude, basal (non-pulsatile) GH secretion, and contribution of basal secretion to total GH secretion in virgin WT (n = 6), pregnant WT (n = 5), virgin Ghrhrlit/lit (n = 6), and pregnant Ghrhrlit/lit(n = 3) mice. The pregnant mice were at gestational ages 14-17 days. *, P < 0.05; **, P < 0.01; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Pregnancy differentially modulates hepatic GHR signaling in WT and Ghrhrlit/lit mice

GH action in the liver plays a key role in controlling growth because GHR signaling stimulates hepatic insulin-like growth factor 1 (IGF-1) secretion, particularly by activating the signal transducer and activator of transcription 5b (STAT5b) pathway (Udy et al., 1997; Davey et al., 2001). GH action in the liver is not only controlled by the pattern of GH secretion but also by hepatic GH sensitivity (Waxman and O’Connor, 2006; de Sousa et al., 2025b). A reduced expression of STAT5b protein was observed in the liver of pregnant WT and Ghrhrlit/lit mice (Figure 4A,C). Phosphorylation of STAT5 (pSTAT5) was also reduced in the liver of pregnant WT compared to virgin animals (Figure 4B-C). Hepatic pSTAT5 was reduced in Ghrhrlit/litmice, compared to virgin WT, and did not change in pregnant animals (Figure 4B-C). Serum IGF-1 and hepatic Igf1 mRNA levels were reduced in pregnant WT mice compared with virgin animals (Figure 4D-E). In contrast, pregnancy increased serum IGF-1 and hepatic Igf1 mRNA levels in Ghrhrlit/lit mice (Figure 4D-E). Corroborating the results of the protein expression, pregnancy reduced Stat5b mRNA levels in both WT and Ghrhrlit/lit mice (Figure 4F). Conversely, hepatic Ghr and Prlr mRNA levels increased during pregnancy in WT and Ghrhrlit/lit mice (Figure 4G-H). Esr1 mRNA expression was reduced in Ghrhrlit/lit mice and was not affected by pregnancy (Figure 4I). Hepatic Socs2 and Cish RNA levels, transcripts that are indicative of GH action and regulate GH sensitivity (Greenhalgh and Alexander, 2004), were also reduced in Ghrhrlit/litmice compared to WT animals (Figure 4J-K). Moreover, pregnancy increased Socs2 and Cish RNA expression only in Ghrhrlit/litmice (Figure 4J-K). Altogether, although WT mice exhibit increased GH secretion and hepatic Ghr expression, molecular analyses do not indicate higher GH action in the liver. In contrast, Ghrhrlit/lit mice showed evidence of greater GH action in the liver, including increased circulating and hepatic IGF-1 levels, as well as greater Socs2 and Cish expression.

Pregnancy differentially modulates hepatic GHR signaling in WT and Ghrhrlit/lit mice.

(A-C) Western blot analysis to determine the hepatic expression of STAT5b and pSTAT5 proteins in virgin WT (n = 6), pregnant WT (n = 6), virgin Ghrhrlit/lit (n = 6), and pregnant Ghrhrlit/lit (n = 6) mice. (D-K) Serum IGF-1 levels and hepatic mRNA expression in virgin WT (n = 11), pregnant WT (n = 10), virgin Ghrhrlit/lit (n = 8), and pregnant Ghrhrlit/lit(n = 7) mice. The pregnant mice were at gestational ages 14-17 days. *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Pregnancy-induced body growth is preserved despite disruption of GH-, ghrelin-, and estrogen-related signaling pathways

To gain further insights into the mechanisms driving post-pregnancy body growth, the effects of pregnancy were investigated across different mouse models. Mice carrying a GHR deletion in hepatocytes (AlbΔGHR) show blunted circulating IGF-1 levels and growth deficits despite higher GH secretion (List et al., 2014; Wasinski et al., 2023; de Sousa et al., 2025b). We observed that a first pregnancy caused a robust increase in body weight and lean mass in AlbΔGHR mice, which was comparable to that seen in control animals (Figure 5A-B). AlbΔGHR females also showed increased body weight after the second pregnancy, but this was explained by pregnancy-induced increases in fat mass (Figure 5C). Thus, GHR signaling in hepatocytes is not necessary to induce pregnancy-induced growth. Ghrelin is a GH secretagogue, and impaired ghrelin signaling may affect body growth (Peris-Sampedro et al., 2021; Punt et al., 2025). However, ghrelin receptor-deficient mice (GHSR KO) showed increases in body weight and lean mass similar to those of control animals (Figure 5D-E). Sex hormones also regulate growth, and estrogen receptor α (ERα) signaling activates the STAT5 pathway in the liver and IGF-1 synthesis (Venken et al., 2005). In addition, estradiol treatment stimulates growth in GHR-deficient mice (Venken et al., 2005), and estrogen response elements are found in the mouse Stat5b gene (Furigo et al., 2014). Nonetheless, hepatocyte-specific ERα ablation did not prevent pregnancy-induced increases in body weight, lean mass, and adiposity (Figure 5F-H). We also evaluated mice carrying brain GHR inactivation, as this is a model of gigantism caused by loss of GH negative feedback (Furigo et al., 2019; Teixeira et al., 2019a; Wasinski et al., 2020; de Sousa et al., 2025a). First pregnancy increased body, lean, and fat mass in NestinΔGHR females, as observed in control animals (Figure 5I-K).

Pregnancy-induced body growth is preserved despite disruption of GH-, ghrelin-, and estrogen-related signaling pathways.

(A-C) The effects of first and second pregnancies on body weight, lean mass, and fat mass in control virgin (n = 10), control 1st pregnancy (n = 7), control second pregnancy (n = 5), AlbΔGHR virgin (n = 17), AlbΔGHR 1st pregnancy (n = 9), and AlbΔGHR second pregnancy (n = 6). (D-E) Body weight and lean mass in control virgin (n = 5), control 1st pregnancy (n = 14), GHSR KO virgin (n = 4), and GHSR KO 1st pregnancy (n = 13). (F-H) Body weight, lean mass, and fat mass in control virgin (n = 7), control 1st pregnancy (n = 4), AlbΔERα virgin (n = 7), and AlbΔERα 1st pregnancy (n = 4). (I-K) The effects of first and second pregnancies on body weight, lean mass, and fat mass in control virgin (n = 11), control 1st pregnancy (n = 11), control second pregnancy (n = 6), NestinΔGHR virgin (n = 6), NestinΔGHR 1st pregnancy (n = 6), and NestinΔGHR second pregnancy (n = 3). *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001. Statistical analysis was performed using two-way ANOVA and Holm-Sidak’s multiple comparisons test.

Case reports of Itabaianinha women with isolated GH deficiency suggest post-pregnancy growth

Growth plate physiology differs between humans and rodents: growth plates remain open after sexual maturation in rodents, whereas they fuse at the end of puberty in humans, thereby terminating longitudinal bone growth (Emons et al., 2011). Although pregnancy-induced longitudinal growth is not observed in normal adult women, we speculate that a slight pregnancy-induced growth may happen in certain special cases. To provide proof-of-concept evidence that gestational factors permanently influence the maternal organism in humans, we studied a cohort of individuals with severe isolated GH deficiency (IGHD) caused by a homozygous loss-of-function mutation (c.57+1G→A) in the GHRHR gene. These individuals are from the Brazilian city of Itabaianinha (Salvatori et al., 1999; Aguiar-Oliveira and Salvatori, 2021), and their mutation replicates effects seen in Ghrhrlit/lit mice (Aguiar-Oliveira and Bartke, 2019).

Using data from this rare cohort, we compared the height of 8 nulliparous women and 9 subjects with one or more pregnancies, both groups with IGHD without lifetime GH replacement therapy. The average number of children per woman was 1.8 children (Aguiar-Oliveira and Salvatori, 2021). Nulliparous women were 7 cm shorter than those with one or more pregnancies, 1.13 ± 0.04 m vs 1.20 ± 0.05 m (P = 0.009; Figure 6A). Moreover, the three shortest women, with heights of 1.07, 1.08, and 1.11 m, were nulliparous. Anecdotal reports from 6 IGHD cases also suggest residual growth in 4 women after pregnancy (Table 1). This information was obtained through face-to-face personal interviews with one of the researchers (M.H. A.-O.), who was also responsible for measuring height before and after pregnancy. Subject 1 reported clear and consistent signs of somatic growth following two pregnancies. She is married to an IGHD individual, and both of her sons (currently 30 and 33 years old) are affected, as expected for this genetic background. According to documented measurements, her height increased from 1.16 m before pregnancy to 1.20 m after her two pregnancies. She also reported an increase in shoe size from 25 to 28 (Brazilian size) and changes in ring size, as a ring previously worn on the fourth finger no longer fit after childbirth. Together, these findings are consistent with clinically perceptible post-pregnancy somatic growth.

Evidence of post-pregnancy growth in women with isolated GH deficiency.

(A) In a cohort of women with isolated GH deficiency caused by a loss-of-function mutation in the GHRHR gene, nulliparous women (n = 8) were shorter than those with one or more pregnancies (n = 9). **, P = 0.009 (unpaired two-tailed Student’s t-test). (B) Comparison of foot size between two sisters with IGHD, one nulliparous and another primiparous. Image of the feet of two sisters with severe isolated GH deficiency caused by a homozygous loss-of-function mutation in the GHRHR gene. In addition to a larger shoe size, subject 2, who had a heterozygous son 9 years ago, is also taller than her nulliparous sister.

Three additional women with IGHD also reported body changes suggestive of growth after pregnancy, though less noticeable than those of subject 1. Subject 2 (43 years old) had a documented increase in height from 1.26 m before pregnancy to 1.27 m at the present. She has one heterozygous son (8.8 years old). She reported an increase in shoe size from 29 to 30, and her graduation ring no longer fits; however, she also gained weight, which may have contributed to the ring size changes. Her feet are larger than her sister’s; she is 44 years old, 1.21 m tall, shoe size 27, and nulliparous (Figure 6B). Subject 3, treated with GH in childhood and early adolescence, was 1.44 m tall before pregnancy. She had a heterozygous boy 13 years ago; now she is 36 years old and 1.47 m tall. Subject 4, currently 28 years old, was treated with GH in childhood and early adolescence. She had a heterozygous girl 3 years ago. Her current height is 1.42 m, but before pregnancy, she was 1.40 m tall. In contrast, 2 sisters with IGHD reported no signs of pregnancy-induced growth. Subject 5, 1.24 m tall, aged 56, has a daughter, aged 35, and two sons, aged 30 and 29, each heterozygous. Subject 6, 1.27 m tall, aged 68, has two heterozygous daughters, aged 46 and 44, respectively. However, no information about their heights before pregnancy is available. Despite these two cases, our data indicate a slight post-pregnancy growth in women with severe IGHD.

Discussion

Pregnancy causes profound endocrine and metabolic changes that are conserved across mammals. The maternal organism undergoes significant hormonal shifts that not only temporarily alter the body but may also lead to lasting effects. Here, we demonstrated that reproductive experience resulted in increased body growth in WT mice, independent of GH. This growth was mainly observed after the first pregnancy, while subsequent reproductive experiences primarily increased body fat. In humans, puberty triggers the fusion of growth plates, leading to the end of longitudinal bone growth (Emons et al., 2011). Height gain is typically limited after menarche, averaging 6.6-8.94 cm depending on the studied cohort (Spear, 2002; Gardstedt-Berghog et al., 2024; Gaete et al., 2025; Xu et al., 2026). Previous studies indicate that pregnant women may experience a slight but significant decrease in height due to weight gain, vertebral compression, and increased lordosis (Scholl et al., 1993; Xu et al., 2026). Therefore, using height as a measure of post-pregnancy growth in humans may be unreliable, which may explain why normal gravidas do not show longitudinal growth. However, increased foot size has been reported after pregnancy (Wetz et al., 2006; Dunn et al., 2012). Despite the lack of evidence for post-pregnancy longitudinal growth in healthy women, the current results in mice suggest that pregnancy activates GH-independent growth pathways. In species with open growth plates, like rodents, this could result in actual longitudinal growth, whereas in humans, it may instead lead to tissue remodeling or physical changes that do not increase height but may have lasting effects that are still not well understood. Thus, the mouse model offers mechanistic insight into conserved pregnancy-related anabolic processes that may operate independently of the somatotropic axis.

Multiparous Ghrhrlit/lit mice showed significant gains in weight (+57%) and lean mass (+59%) after pregnancy compared with age-matched Ghrhrlit/lit virgins. In fact, their relative weight and lean mass gains were much higher than those observed in WT mice, which showed ∼13% more body weight and lean mass than WT virgins. Thus, GH-deficient mice are considerably more sensitive to pregnancy-induced growth than WT mice. The profound growth deficit in Ghrhrlit/lit mice likely enhances their sensitivity to pregnancy-induced growth factors. Human data are in accordance with these findings as children with severe GH deficiency (GHD) show a higher response to GH replacement than patients with less severe or questionable GHD (Savage and Bang, 2012; Wit et al., 2019). Accordingly, patients exhibiting the lowest response during GH provocation tests showed the highest responses to GH replacement treatment (Donbaloglu et al., 2023). Regarding the capacity of women with IGHD to present post-pregnancy growth, residual spinal growth is frequently observed after skeletal maturity, particularly in the 1–2 years following the peak of puberty, often averaging just a few millimeters to 1 cm in height (Ghanem and Rizkallah, 2020). The increased responsiveness to growth factors, as well as the late menarche (±17 years) and, therefore, delayed bone maturation (Aguiar-Oliveira and Salvatori, 2021), can also explain why adult women with severe IGHD retained some growth capacity after pregnancy. Importantly, identifying adult women with untreated IGHD who subsequently experienced pregnancy is extremely challenging. GH replacement therapy is typically initiated early in life, and untreated individuals are rare. Therefore, although a limited number of cases were available for study and the human data relied in part on participants’ self-reported information, this sample represents a highly valuable and unique clinical resource.

Unlike WT females, Ghrhrlit/lit mice show no increase in GH secretion during pregnancy, maintaining suppressed levels similar to those of nulliparous Ghrhrlit/lit mice. Thus, these findings revealed that GHRH signaling seems required for pregnancy-induced increases in GH secretion in mice. However, we cannot distinguish whether the lack of pregnancy-induced increases in GH secretion reflects the absence of GHRH signaling per se or developmental defects resulting from GHRH deficiency. More importantly, this result indicates that pregnancy-induced body growth is independent of GH levels. It is important to note that, despite marked species-specific differences in the GH locus (as summarized in the Introduction), pregnancy-induced growth appears to have been evolutionarily conserved. Although this response may be mediated by distinct mechanisms in rodents and humans, the available evidence suggests that it is largely GH-independent in both species.

However, the expression of genes typically induced by GHR signaling, such as Igf1, Socs2, and Cish, was upregulated in the liver of pregnant Ghrhrlit/lit mice compared with virgin animals. The increased circulating IGF-1 levels in Ghrhrlit/litmice may have contributed to their greater relative growth during gestation. In contrast, pregnant WT mice showed decreased circulating IGF-1 levels and hepatic Igf1 mRNA expression compared to virgin animals, suggesting hepatic GH resistance. Indeed, hepatic STAT5/GH signaling desensitization was described in pregnant mice, which involves increased expression of negative regulators of STAT signaling (Miquet et al., 2005), as observed in the current study. These results are consistent with another study that also observed decreased IGF-1 secretion and hepatic Igf1 mRNA expression in mid-pregnant mice (Travers et al., 1990). In humans, IGF-1 levels decrease during the first trimester of gestation (Caron et al., 2010; Persechini et al., 2015), but increase during mid- and late-pregnancy, when they correlate with placental GH levels (Caufriez et al., 1993). Altogether, pregnancy differentially modulates the hepatic expression of genes commonly induced by GHR signaling in WT and Ghrhrlit/lit mice. Thus, pregnancy-induced growth may occur through distinct mechanisms in WT and Ghrhrlit/litmice. Whereas increased circulating IGF-1 levels may contribute to growth in Ghrhrlit/lit mice, pregnancy-induced growth in WT mice may be mediated through liver-derived IGF-1-independent mechanisms. Locally produced IGF-1 in tissues such as the growth plate, bone, and skeletal muscle may be regulated by gestational factors, thereby promoting body growth. Future studies are needed to investigate this possibility.

In an attempt to identify specific hormonal factors involved in pregnancy-induced body growth, mice with hepatocyte-specific deletion of GHR or ERα were generated. Confirming previous findings that pregnancy-induced body growth is independent of GH, hepatic GHR deletion did not prevent increases in body weight and lean mass in primiparous and multiparous mice. Although estradiol treatment can stimulate growth in GHR-deficient mice and activate the STAT5 pathway and IGF-1 synthesis in the liver (Venken et al., 2005), hepatocyte-specific ERα KO mice also exhibited normal pregnancy-induced body growth. Similar findings were observed in GHSR-deficient mice, demonstrating that pregnancy-induced body growth is ghrelin-independent. Finally, a gigantism model (NestinΔGHR mice) was tested to determine whether pregnancy-induced growth can occur in mice that already have increased baseline lean body mass (+16%) compared with control animals. However, pregnancy also increased body and lean mass in NestinΔGHR mice, in a manner comparable to control animals, suggesting that this growth potential persists even in animals that begin gestation with greater somatic length.

Several questions arose from our findings. For example, why does body growth occur primarily after the first pregnancy, regardless of the age at which it occurs? We speculate that pregnancy hormones, particularly sex hormones, may modify the growth plate of long bones, limiting further growth in subsequent gestations. Future studies should investigate the impact of pregnancy on growth plate structure and function. Another interesting finding is that multiple gestations are associated with increased body adiposity in mice. Leptin resistance is a hallmark of pregnancy, favoring physiological increases in food intake and fat mass (Mounzih et al., 1998; Trujillo et al., 2011; Ladyman et al., 2012; Zampieri et al., 2015; Andreoli et al., 2019; Gustafson et al., 2019). If leptin resistance persists in the postpartum period, it may favor fat retention. However, previous studies indicate normal leptin sensitivity in female mice following pregnancy (Ladyman et al., 2018; Andreoli et al., 2019; Teixeira et al., 2019b). It is also possible that each pregnancy may cause some degree of fat retention, so multiple pregnancies lead to a significant increase in body adiposity simply through an additive effect. Regardless of the mechanism that causes pregnancy-induced fat gain, it is evident that pregnancy may represent a risk factor for obesity (Scholl et al., 1995; Rooney et al., 2005; Amorim et al., 2007).

Our study was not intended to distinguish the effects of pregnancy from those of lactation. Lactation stimulates the secretion of prolactin and oxytocin (Augustine et al., 2008; Ladyman et al., 2010). In mice, lactation is a period of high metabolic demand due to the cost of milk production, which is compensated for by hyperphagia (Woodside et al., 2012; Zampieri et al., 2015; Teixeira et al., 2019a). Thus, lactation may have contributed to the growth of the females and should be considered alongside gestation when interpreting our results. Moreover, female mice housed with sterile males exhibit increased body weight and reduced lifespan (Garratt et al., 2020). Therefore, mating per se, even in the absence of fertilization, is sufficient to influence body weight and longevity, representing an additional factor—beyond pregnancy and lactation—that may have contributed to the parameters analyzed.

Since circulating GH and IGF-1 levels do not fully explain pregnancy-induced growth, several additional endocrine factors may contribute to this phenomenon, each potentially accounting for part of the observed response. Prolactin and placental lactogens (chorionic somatomammotropins in humans) are particularly attractive candidates because they activate intracellular signaling pathways that overlap with those of GH, stimulate cell proliferation, play central roles in maternal physiological adaptations, are markedly elevated during pregnancy, and act on multiple target tissues (Augustine et al., 2008; Ladyman et al., 2010). Therefore, these hormones may contribute to the increased body growth observed in reproductively experienced females. IGF-2 represents another potential mediator. Although best known for its critical role in fetal growth, IGF-2 is abundantly produced at the maternal-fetal interface (Lopez-Tello et al., 2023). While its role in pregnancy-induced maternal growth remains unknown, mice with persistent postnatal IGF-2 expression exhibit increased body growth despite normal circulating IGF-1 concentrations (Younis et al., 2026). Moreover, IGF-2 contributes to maternal adaptations during pregnancy, including the development of mild insulin resistance (Lopez-Tello et al., 2023), making it an attractive candidate to mediate pregnancy-induced body growth. Finally, insulin is a well-established anabolic hormone whose secretion increases during pregnancy (Augustine et al., 2008; Ladyman et al., 2010). Thus, enhanced insulin receptor signaling may also directly or indirectly contribute to pregnancy-induced maternal growth. Investigating the contribution of these factors to the long-term consequences of reproductive experience, particularly the body growth observed in the present study, represents an important direction for future research.

In conclusion, reproductive experience, including mating, pregnancy, and lactation, activates redundant, multi-hormonal anabolic programs involving growth factors and placental lactogens, among others. Our findings indicate that this coordinated endocrine-metabolic environment is sufficient to promote longitudinal skeletal growth, even in the absence of pituitary GH, in species with open growth plates. We also provided evidence that post-pregnancy growth can occur in women with severe IGHD. Thus, reproductive experience has many effects on the maternal organism. Despite the lack of longitudinal growth in normal women, their bodies undergo significant and potentially permanent changes that need further study to understand the long-term impact on women’s health.

Material and Methods

Animals

In the current study, the following mouse strains were used: C57BL/6J (The Jackson Laboratory, Bar Harbor, ME; RRID:IMSR_JAX:000664), Ghrhrlit/lit(The Jackson Laboratory; RRID:IMSR_JAX:000533), AlbCre (The Jackson Laboratory; RRID:IMSR_JAX:003574), Ghrflox (List et al., 2014), ERαflox (The Jackson Laboratory; RRID:IMSR_JAX:032173), Nestincre (The Jackson Laboratory; RRID:IMSR_JAX:003771), and GHSR-deficient mice (Zigman et al., 2005). The generation, genotyping, and validation of the experimental mice were carried out as described in previous studies from our group (Frazao et al., 2013; Furigo et al., 2019; Teixeira et al., 2019a; Wasinski et al., 2020; Wasinski et al., 2022; Wasinski et al., 2023; de Sousa et al., 2025a; de Sousa et al., 2025b). Mice had ad libitum access to regular rodent chow (2.99 kcal/g, 9.4% kcal from fat, 236 g protein/kg; Nuvilab CR-1, Quimtia, Brazil) and filtered water. The experimental procedures were approved by the Ethics Committee on the Use of Animals of the Institute of Biomedical Sciences at the University of São Paulo.

Evaluation of pregnancy-induced growth in mice

The body composition (lean and fat mass) of adult (∼10-week-old) female mice was determined by time-domain nuclear magnetic resonance using the LF50 body composition mice analyzer (Bruker, Germany). Subsequently, the females were mated with sexually experienced males. Upon detection of pregnancy, females were placed in individual cages and maintained with their litters for 3-4 weeks. After weaning, the dams were regrouped in cages with up to five females. Body composition was reassessed 2 weeks after the last contact with the pups (∼20-week-old), which we consider the lactation recovery period. At the same time, body composition was analyzed in age-matched virgin females as a control group. In some experiments, the primiparous females were allowed to mate again, and the entire procedure was repeated to determine the effects of the second pregnancy and lactation cycle. After the last body composition assessment, WT and Ghrhrlit/lit mice were euthanized in a CO2 chamber, and naso-anal length was measured, as were the masses of the liver, heart, kidneys, and brain. Figure 7 summarizes the experimental timeline. Specifically, for GHSR KO mice and their respective controls, lean mass was calculated as total body weight minus fat mass (the sum of periuterine, periovarian, retroperitoneal, inguinal, posterior subcutaneous, and anterior subcutaneous fat depots).

Experimental timeline illustrating the study design.

The schematic was created with the assistance of OpenAI image-generation tools and subsequently reviewed and edited by the authors for scientific accuracy.

Evaluation of pulsatile GH secretion and IGF-1 levels

To adapt the mice to the tail-tip blood sampling procedure, females were acclimated daily for 2-3 weeks before being placed in breeding. This adaptation continued while the females were housed with males. In pregnant females (gestational age 14-17 days) and virgin controls, a total of 36 blood samples were taken from the tail tip every 10 minutes. The procedure initially involved clipping a small portion (1 mm) from the tail tip with a surgical blade. Each blood sample (5 μL for WT and 2 μL for Ghrhrlit/lit) was then transferred to a tube containing PBS with 0.05% Tween-20. After each draw, gentle fingertip pressure was applied to the tail to halt bleeding. The mice continued to move freely within their home cages, with unlimited access to food and water. Immediately after collection, samples were placed on dry ice and stored at -80°C. An enzyme-linked immunosorbent assay (ELISA) was performed to measure blood GH levels, using a 1:22 final dilution, and the following antisera: monkey anti-rat GH as the capture antibody (1:50,000; NHPP, Cat# AFP411S; RRID:AB_2665564) and rabbit anti-rat GH as the detection antibody (1:100,000; NHPP, Cat# AFP5672099, RRID:AB_2721132). GH pulses were detected using the DynPeak pulse detection algorithm’s default settings (Vidal et al., 2012). More details about this ELISA and the analysis of the GH secretion pattern are found in previous publications (Wasinski et al., 2020; Wasinski et al., 2022; de Sousa et al., 2024; de Sousa et al., 2025b). Serum IGF-1 levels were measured with a commercial ELISA kit (#MG100; RRID:AB_2827989; R&D Systems, Minneapolis, MN, USA).

Western Blot

Liver proteins were extracted using RIPA buffer (Sigma-Aldrich, Cat# R0278), containing phosphatase (Sigma-Aldrich, Cat# P5726 and P0044) and protease (Sigma-Aldrich, Cat# P8340) inhibitors. Samples were fractionated by polyacrylamide gel electrophoresis containing SDS (SDS-PAGE) using a running buffer (25 mM Tris; 192 mM Glycine; 0.1% SDS). Then, proteins were transferred to a nitrocellulose membrane overnight at 15 V using transfer buffer (25 mM Tris; 192 mM Glycine; 20% Methanol). Membranes were blocked at room temperature for 2 h in 5% skim milk diluted in TBS-T buffer (10 mM Tris–HCl; 150 mM NaCl; 0.05% Tween-20). Membranes were incubated overnight in primary antibodies diluted in TBS-T containing 3% albumin. Membranes were washed three times for 10 min with TBS-T and incubated at room temperature for 1 h with respective secondary antibodies, diluted in 5% skim milk in TBS-T. After three 10-minute washes with TBS-T, the membranes were treated with chemiluminescent reagents (Clarity Western ECL Substrate, Bio-Rad, Hercules, CA; Cat# 170-5060). Bands on the membranes were revealed by automatic imaging systems. The intensities of the bands were quantified by densitometry using ImageJ software (https://imagej.net/ij/) and normalized to total protein expression in the gel (Ponceau staining). Primary antibodies used for blotting were anti-phosphoTyr694-STAT5 antibody (Cell Signaling Technology, Beverly, MA; Cat# 9351; RRID:AB_2315225; 1:1,000), anti-STAT5b antibody (Santa Cruz, Dallas, TX; Cat# sc-377069; 1:1,000), and anti-β-actin (Santa Cruz; Cat# sc-47778, RRID:AB_626632; 1:2,500).

Gene expression

Total RNA from the liver was extracted with TRIzol (Invitrogen, Carlsbad, CA), followed by incubation in DNase I RNase-free (MilliporeSigma, St. Louis, MO, USA). Reverse transcription was performed using 2 µg of total RNA, SuperScript II Reverse Transcriptase (Invitrogen), and random primers p(dN)6 (MilliporeSigma). Real-time PCR was performed using the 7500TM Real-Time PCR System (Applied Biosystems, Warrington, UK) and SYBR Green Gene Expression PCR Master Mix (Applied Biosystems. The following primers were used: Actb (forward: gctccggcatgtgcaaag; reverse: catcacaccctggtgccta), Cish (forward: tcgggaatctgggtggtact; reverse: taggaatgtaccctccggca), Esr1 (forward: gcagatagggagctggttca; reverse: tggagattcaagtccccaaa), Ghr (forward: atcaatccaagcctggggac; reverse: acagctgaatagatcctgggg), Igf1 (forward: ccacactgacatgcccaaga; reverse: gtacttcctttccttctcctttgc), Prlr (forward: cagtaaatgccaccaacgaa; reverse: gaggaggctctggttcaaca), Ppia (forward: tatctgcactgccaagactgagt; reverse: cttcttgctggtcttgccattcc), Socs2 (forward: cgcgagctcagtcaaacag; reverse: aagttccttctggagcctctt), and Stat5b (forward: ggactccgtccttgataccg; reverse: tccatcgtgtcttccagatcg). Data were normalized to the Actb (β-actin) and Ppia expression. Relative quantification of mRNA was calculated by 2-ΔΔCt.

Statistical analysis

When comparing two groups, the data were analyzed using a two-tailed, unpaired Student’s t-test. When comparing three groups simultaneously, one-way ANOVA followed by the Newman-Keuls multiple comparisons test was used. Two-way ANOVA and Holm–Sidak’s multiple comparisons test were used to analyze two variables (pregnancy and mutation effects). Statistical analyses were conducted with Prism software (GraphPad, San Diego, CA). All results are presented as mean ± standard error of the mean.